---
title: "Session 2 — Disease Mapping and Spatial Patterns"
subtitle: "Choropleth classification · extreme value maps · counts vs rates · variance instability · SMR · Empirical Bayes smoothing · spatial weights · Moran's I · LISA"
engine: knitr
---

```{=html}
<script>document.body.classList.add("block-esda");</script>
```

```{r}
#| include: false
source("_helpers.R")
```

Session 2 moves from **making maps** to **evaluating spatial patterns**. It has two halves.

The first half is about maps of disease. How you classify a variable into colours changes what the reader sees, and **crude rates from small populations are unstable**: an area with 50 inhabitants can show a very high or very low rate purely by chance. The standardized mortality ratio (SMR) and Empirical Bayes smoothing are the first tools for this.

The second half formalises Tobler's law. To ask whether nearby areas are similar, you first have to define **who is a neighbour** (a spatial weights matrix). Then **Moran's I** measures global spatial autocorrelation, and **local Moran's I (LISA)** says *where* the clusters and outliers are.

::: {.callout-note collapse="true"}
## Sources and session
**Slides:** "session_2_v0.pdf" (Qinyun Lin, 6 October 2026, 15:15–17:00). Several figures on the slides come from the GeoDa workbook (geodacenter.github.io/workbook) and Anselin's course material.

**Learning objectives (Canvas post):** after this session you should be able to

1. explain why mapping counts and crude rates can produce different spatial patterns;
2. explain why estimates from areas with small populations can be unstable;
3. describe different ways of defining spatial neighbours;
4. interpret positive and negative spatial autocorrelation;
5. explain the basic idea of Moran's I;
6. distinguish between global and local spatial autocorrelation.

Topics that match an objective carry 🎯.
:::

::: {.callout-tip collapse="true"}
## Before you start: three ideas from earlier courses
**Quartiles and the IQR.** Q1 and Q3 cut off the lowest and highest 25%. IQR = Q3 − Q1. A box plot calls values beyond Q1 − 1.5·IQR or Q3 + 1.5·IQR outliers.

**Variance of a proportion.** If a proportion is estimated from n people with true probability π, its variance is π(1 − π)/n. Halving n doubles the variance.

**What a p-value means.** The probability, **if the null hypothesis were true**, of a test statistic at least as extreme as the one observed. It is not the probability that the null hypothesis is true. The slides ask you to recall this before the Moran's I test.
:::

## 2.1 Thematic (choropleth) maps and classification methods {#t2-1 .topic flag="inferred"}

A **thematic** or **choropleth map** shows values with colours or patterns inside geographic areas. Two choices change the picture: the **classification method** (where the class boundaries go) and the **number of classes**.

| Method | How the class breaks are chosen | Strength | Weakness |
|---|---|---|---|
| **Quantile** | Each class has the **same number of areas** (3 = tertiles, 4 = quartiles, 5 = quintiles) | Every colour is used equally; good for ranking | Very different values can share a class; close values can be split |
| **Natural breaks (Jenks)** | Breaks **minimise variation within** classes and **maximise differences between** them | Follows gaps in the data | Classes can hold very unequal numbers of areas; hard to compare maps |
| **Equal intervals** | The range is cut into classes of equal width | Easy to read | One extreme value can push most areas into one class |
| **Manual intervals** | You choose meaningful cut-offs (for example clinical thresholds) | Interpretable | Needs justification |
| **Standard deviation** | Values are expressed in SD units from the mean | Shows how far from average; spots outliers | Assumes a roughly symmetric distribution |

```{r}
#| label: fig-classification
#| echo: false
#| fig-width: 9
#| fig-height: 3.6
#| fig-cap: "The same skewed rates on a 10 × 10 grid with three classification methods and five classes. Quantiles fill every class equally and make the map look busy; equal intervals put most areas in the lowest class; natural breaks follow gaps in the values. Same data, three impressions. (Natural breaks approximated here with one-dimensional k-means.)"
side <- 10
set.seed(21)
# skewed rates: most areas low, a few very high
rate <- round(exp(1.2 + 0.9 * smooth_surface(side, 4, seed = 5) * 0.6 + rnorm(side^2, 0, 0.35)), 1)
cols <- seq_cols(5)
quantile_br <- quantile(rate, seq(0, 1, 0.2))
equal_br <- seq(min(rate), max(rate), length.out = 6)
# natural breaks approximated by 1-D k-means on the values
km <- kmeans(rate, centers = quantile(rate, c(0.1, 0.3, 0.5, 0.7, 0.95)), iter.max = 50)
ord <- order(km$centers)
natural_br <- c(min(rate), sort(tapply(rate, km$cluster, max))[1:4], max(rate))
par(mfrow = c(1, 3), mar = c(4.5, 0.5, 2.5, 0.5))
for (m in list(list(quantile_br, "Quantile (quintiles)"), list(equal_br, "Equal intervals"),
               list(natural_br, "Natural breaks"))) {
  draw_grid(rate, side, cols, m[[1]], main = m[[2]])
  counts <- table(cut(rate, m[[1]], include.lowest = TRUE))
  mtext(paste("areas per class:", paste(counts, collapse = " · ")), side = 1, line = 0.3, cex = 0.7)
  mtext(paste("breaks:", paste(round(m[[1]], 1), collapse = ", ")), side = 1, line = 1.4, cex = 0.62)
}
```

**There is no universally best method.** It depends on the distribution of the data, the purpose of the map and the comparison you want the reader to make.

:::: {.quiz}
Your variable is very skewed: most municipalities have low rates and a few have extreme values. With equal intervals and 5 classes, what will the map look like?

- **Almost every area in the lowest class, a few in the top ones**
- Each class holding the same number of areas, as in quintiles
- Classes chosen at the natural gaps between groups of values
- All areas coloured by how many SDs they lie from the mean

::: {.exp}
Equal intervals split the *range* evenly. A few extreme values stretch the range, so most areas fall in the bottom interval. A quantile map would instead force equal counts per class.
:::
::::

## 2.2 Extreme value maps: percentile, box and standard deviation maps {#t2-2 .topic flag="keyq"}

Extreme value maps are designed to **highlight outliers** rather than show the whole distribution evenly.

| Map | Classes |
|---|---|
| **Percentile map** | Fixed percentile cut-offs: < 1%, 1–10%, 10–50%, 50–90%, 90–99%, > 99% |
| **Box map** | The box plot's categories: lower outlier, < 25%, 25–50%, 50–75%, > 75%, upper outlier |
| **Standard deviation map** | Classes in SD units around the mean |

**Box map fences.** Lower bound LB = Q1 − hinge × IQR, upper bound UB = Q3 + hinge × IQR.

::: {.worked}
| | Hinge = 1.5 | Hinge = 3 |
|---|---|---|
| Q1, Q3, IQR (rent example on the slide) | 1000, 1362.5, 362.5 | same |
| LB | 1000 − 1.5 × 362.5 = **456.25** | 1000 − 3 × 362.5 = **−87.5** |
| UB | 1362.5 + 1.5 × 362.5 = **1906.25** | 1362.5 + 3 × 362.5 = **2450** |
| Effect | More outliers flagged | **Fewer and more extreme** outliers flagged |
:::

**The slide's question: which map is the box map and which is the percentile map?** Read the legend. A box map's legend has "Lower outlier", "< 25%", "25–50%", "50–75%", "> 75%", "Upper outlier", with the outlier classes holding however many areas lie beyond the fences (possibly none). A percentile map's legend has the fixed bins < 1%, 1–10%, 10–50%, 50–90%, 90–99%, > 99%, so its extreme classes always hold about 1% of areas each.

<div class="drill" data-drill="boxmap"></div>

## 2.3 Mapping categorical variables and co-location {#t2-3 .topic}

A **unique value map** gives each category its own colour (for example urban / suburban / rural).

A **co-location map** extends it to show where **several characteristics overlap**. The slide's example marked US counties with both **persistently high COVID-19 mortality** and a **high population share of a particular racial/ethnic group**, so you can see where the two coincide. Co-location shows overlap, not cause.

## 2.4 Counts vs rates, and the problem of variance instability {#t2-4 .topic flag="objective"}

Two kinds of variables behave differently on a map:

| | Spatially **extensive** | Spatially **intensive** |
|---|---|---|
| Depends on population size? | **Yes** | No, standardized by population |
| Example | Number of cases or deaths | Rate or risk: cases per population |
| What a map shows | Mostly **where people live** | Where the **risk** is high |

Mapping **counts** therefore mostly reproduces a population map: big cities have the most cases. **Crude rates** fix this, but bring a new problem.

**Variance instability.** For area i with observed cases Oᵢ and population at risk Pᵢ, the crude risk is rᵢ = Oᵢ / Pᵢ. If the underlying risk is πᵢ,

$$
\operatorname{Var}(r_i) = \frac{\pi_i (1 - \pi_i)}{P_i}.
$$

Small Pᵢ → large variance → unstable crude estimates. **Areas with small populations show very high or very low crude rates simply because of sampling variability.** This motivates smoothing.

**The slide's question:** with the same underlying πᵢ, which area has the more variable observed risk, Pᵢ = 100 or Pᵢ = 10,000? **Pᵢ = 100**: its variance is 100 times larger (its standard deviation 10 times larger).

```{r}
#| label: funnel
#| echo: false
#| fig-width: 9
#| fig-height: 3.6
#| fig-cap: "Simulated: 300 areas that all share the same true risk of 1%. Left: crude rates against population size. The most extreme rates, both high and low, all belong to small areas, a 'funnel'. Right: the map-maker's view, the 15 highest crude rates; nearly all are small areas with no real excess risk."
set.seed(2)
true_risk <- 0.01
population <- round(exp(runif(300, log(80), log(30000))))
cases <- rbinom(300, population, true_risk)
crude_rate <- 1000 * cases / population
par(mfrow = c(1, 2), mar = c(4.2, 4.2, 2.5, 1))
plot(population, crude_rate, log = "x", pch = 19, col = "#2f5d8a80",
     xlab = "population at risk (log scale)", ylab = "crude rate per 1,000",
     main = "Same true risk everywhere")
abline(h = 1000 * true_risk, lty = 2)
upper <- 1000 * (true_risk + 1.96 * sqrt(true_risk * (1 - true_risk) / sort(population)))
lower <- 1000 * pmax(0, true_risk - 1.96 * sqrt(true_risk * (1 - true_risk) / sort(population)))
lines(sort(population), upper, col = "#c92a2a"); lines(sort(population), lower, col = "#c92a2a")
top15 <- order(crude_rate, decreasing = TRUE)[1:15]
barplot(population[top15], names.arg = round(crude_rate[top15], 1), las = 2, log = "y",
        col = "#d9480f", ylab = "population (log scale)", xlab = "crude rate per 1,000",
        main = "The 15 'highest-risk' areas")
```

:::: {.quiz}
A national map of crude cancer mortality shows that the very highest **and** the very lowest rates are almost all in sparsely populated rural municipalities. What is the most likely explanation?

- **Small populations give unstable rates that scatter in both directions**
- Rural areas have a genuinely more extreme mix of cancer risk factors
- Counts were mapped instead of rates, so population size dominates
- The classification method put rural areas in the extreme classes

::: {.exp}
Extremes in both directions in small areas is the signature of variance instability. A real rural risk factor would push rates one way, not both ways.
:::
::::

## 2.5 Comparing areas with the SMR (excess risk) {#t2-5 .topic flag="keyq"}

An **excess risk map** compares each area's rate with a **reference risk** (national or regional). The reference risk is estimated from **all observations pooled**:

$$
\hat\pi = \frac{\sum_{i=1}^{n} O_i}{\sum_{i=1}^{n} P_i} = \frac{\text{total events}}{\text{total population at risk}}.
$$

Expected count and standardized mortality ratio (SMR):

$$
E_i = \hat\pi \times P_i, \qquad \text{SMR}_i = \frac{O_i}{E_i} = \frac{O_i}{\hat\pi P_i} = \frac{r_i}{\hat\pi}.
$$

**The slide's two questions:**

1. *Is π̂ equal to the average of all rates?* **No.** π̂ is a **population-weighted** average: total events over total population. The simple mean of the area rates gives every area the same weight, so a few small areas with extreme rates can pull it far away. They coincide only if all areas have the same population.
2. *How to interpret SMRᵢ > 1?* Area i had **more events than expected** if the reference risk applied there. SMR = 1.3 means 30% more than expected. It is not, on its own, evidence of a real excess: a small Eᵢ makes the SMR unstable (Session 4).

::: {.worked}
**Slide practice.** Overall cancer risk 10%. Area A has 1,000 residents at risk; 20 get cancer.

| Approach | Calculation | SMR |
|---|---|---|
| 1: expected count | E = 1000 × 10% = 100; SMR = 20 / 100 | **0.2** |
| 2: rate ratio | crude rate = 20 / 1000 = 2%; SMR = 2% / 10% | **0.2** |

Area A has 80% fewer cases than expected.
:::

Expected counts can also use **stratum-specific** reference risks (by age and sex): indirect standardization, worked through in [Session 4](session4.qmd#t4-3).

<div class="drill" data-drill="smr"></div>

## 2.6 Empirical Bayes smoothing {#t2-6 .topic flag="objective"}

**Idea:** replace each crude rate by a **weighted average** of the area's own crude rate and the overall (reference) rate:

$$
\hat\pi_i^{EB} = w_i\, r_i + (1 - w_i)\, \theta, \qquad w_i = \frac{\sigma^2}{\sigma^2 + \mu / P_i}.
$$

| Symbol | Meaning |
|---|---|
| rᵢ | Crude rate in area i |
| θ (= μ) | Prior estimate: the reference rate (overall average or another standard) |
| σ² | Variance of the **true** rates between areas (how much real variation there is) |
| μ / Pᵢ | Approximately the **sampling** variance of rᵢ: large when Pᵢ is small |
| wᵢ | Weight on the area's own data, between 0 and 1 |

**Reading the weight.** wᵢ compares real between-area variation (σ²) with noise (μ/Pᵢ):

- **large, informative area**: μ/Pᵢ is tiny, wᵢ ≈ 1 → **little shrinkage**, the area keeps its own rate;
- **small, uninformative area**: μ/Pᵢ is large, wᵢ is small → **strong shrinkage** toward the reference rate.

```{r}
#| label: fig-eb-dumbbell
#| echo: false
#| fig-width: 9
#| fig-height: 3.8
#| fig-cap: "Empirical Bayes smoothing, simulated: 40 areas whose true risks vary moderately around 1%. Left: each grey line shows how far a crude rate (orange) moves to its EB estimate (blue). Small areas are pulled almost all the way to the reference rate; large areas barely move. Right: the weight w on the area's own rate grows with population."
set.seed(8)
n_areas <- 40
true_risk <- 0.01 * exp(rnorm(n_areas, 0, 0.2))        # real differences between areas
population <- round(exp(seq(log(150), log(40000), length.out = n_areas)))
cases <- rbinom(n_areas, population, true_risk)
crude <- cases / population
mu <- sum(cases) / sum(population)
# Marshall's moment estimate of the between-area variance sigma^2
weighted_var <- sum(population * (crude - mu)^2) / sum(population)
between_var <- max(weighted_var - mu / mean(population), 1e-7)
w <- between_var / (between_var + mu / population)
eb <- w * crude + (1 - w) * mu
par(mfrow = c(1, 2), mar = c(4.2, 4.2, 2.5, 1))
plot(population, 1000 * crude, log = "x", pch = 19, col = col_orange, ylim = range(1000 * c(crude, eb)),
     xlab = "population (log scale)", ylab = "rate per 1,000", main = "Crude rate → EB estimate")
arrows(population, 1000 * crude, population, 1000 * eb, length = 0.05, col = col_grey)
points(population, 1000 * eb, pch = 19, col = col_blue)
abline(h = 1000 * mu, lty = 2)
legend("topright", c("crude rate", "EB smoothed", "reference rate"), pch = c(19, 19, NA), lty = c(NA, NA, 2),
       col = c(col_orange, col_blue, "black"), bty = "n", cex = 0.8)
plot(population, w, log = "x", pch = 19, col = col_blue, ylim = c(0, 1),
     xlab = "population (log scale)", ylab = "weight on own crude rate, w", main = "The weight grows with population")
abline(h = c(0, 1), col = "#dee2e6")
text(min(population), 0.2, "strong shrinkage", pos = 4, cex = 0.8)
text(max(population), 0.92, "little shrinkage", pos = 2, cex = 0.8)
```

EB smoothing reduces the influence of unstable extreme estimates from small populations.

::: {.misconception}
"Empirical Bayes is a Bayesian method." Not quite. The prior (θ and σ²) is **estimated from the same data** rather than specified beforehand, and the method produces point estimates without a full posterior distribution. It borrows the Bayesian *idea* of shrinkage. (The same trick is used in moderated t-tests for gene expression: small samples get their variance shrunk.) Fully Bayesian disease mapping is Session 4.
:::

**Three maps of the same data** (slides 17–18): a **raw rate map**, an **excess risk map** (SMR) and an **EB smoothed map** can look quite different. In the raw and excess risk box maps, small areas often appear as outliers; after EB smoothing, many of them move back toward the middle and the remaining outliers are more trustworthy.

<div class="drill" data-drill="eb"></div>

::: {.callout-tip}
## When to use EB smoothing
When areas differ a lot in population size and the goal is a map of risk that is not dominated by small-area noise. It is quick and needs no model fitting.
:::

::: {.callout-warning}
## Limitations
It ignores geography: every area is shrunk toward the same global mean, not toward its neighbours (spatial EB versions exist). It gives no full uncertainty for each area. A genuinely high-risk small area is pulled toward the mean too, so real local excesses can be hidden.
:::

## 2.7 Spatial autocorrelation and the null of spatial randomness {#t2-7 .topic flag="objective"}

**Spatial autocorrelation** formalises Tobler's law: the correlation of a variable **with itself across space**.

```{r}
#| label: ac-patterns
#| echo: false
#| fig-width: 9
#| fig-height: 3.3
#| fig-cap: "Three 8 × 8 patterns with the same number of black and white cells (as on slide 20, after O'Sullivan and Unwin 2002). Moran's I uses binary rook weights."
rook_weights <- function(side) {
  n <- side^2
  W <- matrix(0, n, n)
  for (i in 1:n) {
    r <- (i - 1) %/% side; c <- (i - 1) %% side
    for (d in list(c(-1, 0), c(1, 0), c(0, -1), c(0, 1))) {
      rr <- r + d[1]; cc <- c + d[2]
      if (rr >= 0 && rr < side && cc >= 0 && cc < side) W[i, rr * side + cc + 1] <- 1
    }
  }
  W
}
morans_i <- function(x, W) {
  z <- x - mean(x)
  (length(x) / sum(W)) * sum(W * outer(z, z)) / sum(z^2)
}
side <- 8
W8 <- rook_weights(side)
halves  <- rep(c(rep(1, 4), rep(0, 4)), times = side)
checker <- as.numeric((rep(0:7, each = 8) + rep(0:7, times = 8)) %% 2 == 0)
set.seed(11); random_pattern <- sample(halves)
par(mfrow = c(1, 3), mar = c(1, 1, 3, 1))
for (p in list(list(halves, "Positive"), list(checker, "Negative"), list(random_pattern, "None (random)"))) {
  image(matrix(p[[1]], side, side), col = c("white", "black"), axes = FALSE,
        main = sprintf("%s: I = %.2f", p[[2]], morans_i(p[[1]], W8)))
  box()
}
```

**Testing.** A test statistic summarises the data and is compared with its distribution **under the null hypothesis**. For spatial autocorrelation the null is **spatial randomness**: the observed values are not systematically associated with their locations. Any arrangement of the values over the map would be equally likely.

That suggests a direct test: **randomly permute (reshuffle) the values over the locations** many times, compute the statistic each time, and see where the observed value falls.

To measure autocorrelation you need two kinds of similarity at once:

- **attribute similarity**: are two values alike? (for Moran's I, the product of deviations from the mean, zᵢzⱼ);
- **locational similarity**: are the two areas neighbours? (the spatial weight wᵢⱼ).

## 2.8 Spatial weights: contiguity (rook and queen) {#t2-8 .topic flag="objective"}

Before testing for autocorrelation you must define **neighbours**. A **spatial weights matrix W** records, for every pair of areas, whether (or how strongly) they are neighbours: wᵢⱼ = 1 if i and j are neighbours, 0 otherwise, and wᵢᵢ = 0. Choosing W means thinking about **how far an influence can reach**.

::: {.worked}
**3 × 3 grid, rook contiguity** (from the slide). Areas are numbered row by row:

| | | |
|---|---|---|
| 1 | 2 | 3 |
| 4 | 5 | 6 |
| 7 | 8 | 9 |

| | 1 | 2 | 3 | 4 | 5 | 6 | 7 | 8 | 9 |
|---|---|---|---|---|---|---|---|---|---|
| **1** | 0 | 1 | 0 | 1 | 0 | 0 | 0 | 0 | 0 |
| **2** | 1 | 0 | 1 | 0 | 1 | 0 | 0 | 0 | 0 |
| **3** | 0 | 1 | 0 | 0 | 0 | 1 | 0 | 0 | 0 |
| **4** | 1 | 0 | 0 | 0 | 1 | 0 | 1 | 0 | 0 |
| **5** | 0 | 1 | 0 | 1 | 0 | 1 | 0 | 1 | 0 |
| **6** | 0 | 0 | 1 | 0 | 1 | 0 | 0 | 0 | 1 |
| **7** | 0 | 0 | 0 | 1 | 0 | 0 | 0 | 1 | 0 |
| **8** | 0 | 0 | 0 | 0 | 1 | 0 | 1 | 0 | 1 |
| **9** | 0 | 0 | 0 | 0 | 0 | 1 | 0 | 1 | 0 |

The matrix is symmetric, the diagonal is 0, the centre (5) has 4 neighbours and corners have 2.
:::

| | **Rook** contiguity | **Queen** contiguity |
|---|---|---|
| Neighbours if they share | A **boundary** (edge), not just a corner | A boundary **or a corner** |
| Number of neighbours | Fewer | **More** (includes all rook neighbours) |
| Higher orders | Yes | Yes |
| Islands / unconnected areas | Not connected | Not connected |

```{r}
#| label: fig-neighbour-types
#| echo: false
#| fig-width: 9
#| fig-height: 3.4
#| fig-cap: "Neighbour definitions on a regular grid. The blue area is the target; red cells are its neighbours. An edge or corner area has fewer neighbours than an interior one."
side <- 7
target <- 25           # row 4, col 4 (interior)
rc <- grid_rc(side)
dr <- abs(rc$row - rc$row[target]); dc <- abs(rc$col - rc$col[target])
cols <- c("white", "#f4a582", "#d6604d", "#2f5d8a")   # none, 2nd order, 1st order, target
show <- function(cls, main, sub) {
  draw_grid(cls, side, cols, border = "#ced4da", main = main)
  mtext(sub, side = 1, line = 0.3, cex = 0.72)
}
par(mfrow = c(1, 4), mar = c(2, 0.5, 2.5, 0.5))
rook <- ifelse(dr + dc == 1, 3, 1); rook[target] <- 4
show(rook, "Rook, 1st order", "4 neighbours: shared edges")
queen <- ifelse(pmax(dr, dc) == 1, 3, 1); queen[target] <- 4
show(queen, "Queen, 1st order", "8 neighbours: edges and corners")
queen2 <- ifelse(pmax(dr, dc) == 1, 3, ifelse(pmax(dr, dc) == 2, 2, 1)); queen2[target] <- 4
show(queen2, "Queen, 2nd order", "dark: 1st order · light: 2nd order ring")
corner <- 1
dr <- abs(rc$row - rc$row[corner]); dc <- abs(rc$col - rc$col[corner])
cq <- ifelse(pmax(dr, dc) == 1, 3, 1); cq[corner] <- 4
show(cq, "Queen, corner area", "only 3 neighbours at the edge")
```

**The slide's question:** if both are first order, which has more neighbours? **Queen**, because it adds the corner (diagonal) neighbours. On a regular grid, an interior cell has 4 rook and 8 queen neighbours.

**Higher order.** Second-order neighbours are **neighbours of neighbours** (the next ring out). Software may or may not include the first order as well; the slide's figure shows the target county, its first-order queen ring and the second-order ring. Same area, different W, different results.

**Row standardization.** Dividing each row by its row sum makes each row sum to 1, so wᵢⱼ = 1/nᵢ. Then Σⱼ wᵢⱼ xⱼ is the **average of the neighbours' values**, the *spatial lag*. This comes back in Moran's scatter plot, the SAR model and the CAR prior.

<div class="drill" data-drill="neighbours"></div>

## 2.9 Spatial weights: distance bands and k-nearest neighbours {#t2-9 .topic flag="objective"}

Distance-based weights use the distance between points (or area centroids).

**Distance metrics**

| Metric | Formula / idea | Use when |
|---|---|---|
| **Euclidean** (straight line) | $d_{ij} = \sqrt{(x_i - x_j)^2 + (y_i - y_j)^2}$ | Coordinates are **projected** (on a plane, in metres) |
| **Great circle / arc distance** | Distance along the Earth's curved surface | **Unprojected** longitude/latitude |

**Distance-band weights.** i and j are neighbours whenever j is within a **critical distance** of i. The result is **symmetric**. To avoid isolates, the threshold is often set to the **largest nearest-neighbour distance** (so every area has at least one neighbour), but then areas in dense clusters can get **very many** neighbours.

**Isolates** are areas with no neighbours. They drop out of all spatial analyses (autocorrelation tests, spatial regression), although they are fine in non-spatial analysis. Ignoring them is a problem for spatially lagged variables and local statistics.

**k-nearest-neighbour (KNN) weights.** Choose k (say 4); each area gets its k nearest neighbours (by centroid distance).

- **Connects every area**: no isolates.
- **Not symmetric**: B can be among A's 4 nearest without A being among B's 4 nearest (B may sit in a dense cluster). It can be made symmetric, for example $w^*_{ij} = (w_{ij} + w_{ji})/2$.
- **Ties**: several areas at the same distance; choose randomly or include all.

| | Contiguity | Distance band | KNN |
|---|---|---|---|
| Symmetric? | Yes | Yes | **No** (unless symmetrized) |
| Isolates possible? | Yes (islands) | Yes, unless threshold ≥ max nearest-neighbour distance | **No** |
| Number of neighbours | Varies | Varies a lot (dense vs sparse areas) | **Exactly k** for every area |

```{r}
#| label: fig-knn-asym
#| echo: false
#| fig-width: 9
#| fig-height: 3.6
#| fig-cap: "The same 15 points with two distance-based definitions. Left: a 2-unit distance band links the dense town heavily and leaves a remote point as an isolate. Right: with k = 3 nearest neighbours every point has exactly three links (arrows), even remote villages, and the links are not symmetric: a village points into town, but the town points do not point back."
set.seed(4)
pts <- rbind(cbind(rnorm(12, 3, 0.6), rnorm(12, 3, 0.6)),    # dense town
             cbind(c(8, 8.9, 9.6), c(7.8, 6.6, 3.2)))         # three remote villages
n <- nrow(pts)
d <- as.matrix(dist(pts))
par(mfrow = c(1, 2), mar = c(1, 1, 2.5, 1))
plot(pts, asp = 1, pch = 19, col = col_blue, axes = FALSE, xlab = "", ylab = "", main = "Distance band (2 units)")
for (i in 1:n) for (j in 1:n) if (i < j && d[i, j] <= 2) segments(pts[i, 1], pts[i, 2], pts[j, 1], pts[j, 2], col = "#2f5d8a66")
points(pts, pch = 19, col = col_blue)
deg <- rowSums(d <= 2) - 1
if (any(deg == 0)) text(pts[deg == 0, 1], pts[deg == 0, 2], "isolate", pos = 2, cex = 0.75, col = col_orange)
mtext("dense town: many links · remote point: none", side = 1, line = -0.5, cex = 0.75)
plot(pts, asp = 1, pch = 19, col = col_blue, axes = FALSE, xlab = "", ylab = "", main = "k-nearest neighbours (k = 3)")
for (i in 1:n) {
  nn <- order(d[i, ])[2:4]
  arrows(pts[i, 1], pts[i, 2], pts[nn, 1], pts[nn, 2], length = 0.06, col = "#d9480f88")
}
points(pts, pch = 19, col = col_blue)
mtext("everyone has 3 links; remote villages reach into town", side = 1, line = -0.5, cex = 0.75)
```

:::: {.quiz}
You study Swedish municipalities, including island municipalities such as Gotland. You need every area to have neighbours for a LISA analysis. Which weights guarantee that?

- **k-nearest-neighbour weights, for example k = 4**
- First-order rook contiguity weights between areas
- First-order queen contiguity weights between areas
- A distance band shorter than the largest nearest-neighbour gap

::: {.exp}
Contiguity weights leave islands without neighbours, and a distance band only connects everyone if its threshold is at least the largest nearest-neighbour distance. KNN always gives each area exactly k neighbours.
:::
::::

## 2.10 Describing a weights matrix: connectivity histogram and graph {#t2-10 .topic flag="keyq"}

Two diagnostics show what a weights matrix looks like:

- **Connectivity histogram**: how many areas have 1, 2, 3, … neighbours.
- **Connectivity graph**: a line between the centroids of every pair of neighbours.

**The slides' questions:**

*What does the histogram look like with KNN weights?* A **single bar** at k: every area has exactly k neighbours. (If you symmetrize the KNN matrix, some areas gain neighbours and the histogram spreads to the right of k.)

*Which graph suggests KNN and which suggests a distance band?* With a **distance band**, dense areas (cities) have **very many** links, sparse areas have few, and isolated points may have **none**. With **KNN**, every point has the same number of links, even remote points, which reach out with long lines to their nearest neighbours.

## 2.11 Global Moran's I {#t2-11 .topic flag="objective"}

$$
I = \frac{\sum_{i=1}^{n}\sum_{j=1}^{n} w_{ij}\, z_i z_j \,/\, S_0}{\sum_{i=1}^{n} z_i^2 \,/\, n},
\qquad z_i = x_i - \bar x, \qquad S_0 = \sum_i\sum_j w_{ij}.
$$

**Reading it.** The numerator is a **cross-product** between each value and its neighbours (attribute similarity × locational similarity), averaged over all neighbour pairs. The denominator is the variance. So Moran's I is like a correlation between a variable and its **spatial lag**.

| zᵢ and its neighbours' zⱼ | Product zᵢzⱼ | Contribution |
|---|---|---|
| Both above the mean (high next to high) | + | Positive |
| Both below the mean (low next to low) | + | Positive |
| One above, one below (high next to low) | − | Negative |

**Inference.** H₀: spatial randomness. Two ways to judge whether the observed I is unusual:

| Approach | How | Pros / cons |
|---|---|---|
| **Analytical** | Normal approximation with the theoretical mean and variance of I | Fast; **sensitive to distributional assumptions** |
| **Permutation** (computational) | Randomly reshuffle the values over the areas many times, compute I each time, compare the observed I with this **reference distribution** → **pseudo p-value** | More **robust**; limited to the actual sample. More permutations (999 rather than 99) give more stable results |

Moran's I **depends on the weights**: the same data with another W give another I.

::: {.misconception}
"Under the null hypothesis Moran's I equals 0." Its expected value under spatial randomness is $E[I] = -1/(n-1)$, slightly negative. For 100 areas that is about −0.01, so in practice "about 0". With very few areas the difference matters.
:::

The pseudo p-value from R permutations is (number of permuted I ≥ observed I, plus 1) / (R + 1). With 999 permutations the smallest possible value is 1/1000 = 0.001.

<div class="drill" data-drill="moran"></div>

**Try it: the permutation test in your browser.** The code builds an 8 × 8 grid with a smooth gradient plus noise, computes Moran's I with rook weights, and compares it with 999 random reshuffles. Change `trend_strength` to 0 and run again.

```{webr-r}
#| label: moran-permutation
side <- 8
n_areas <- side^2
trend_strength <- 1        # 0 = no spatial pattern

# binary rook weights for an 8 x 8 grid
row_of <- (seq_len(n_areas) - 1) %/% side
col_of <- (seq_len(n_areas) - 1) %% side
W <- outer(seq_len(n_areas), seq_len(n_areas), function(i, j)
  as.numeric(abs(row_of[i] - row_of[j]) + abs(col_of[i] - col_of[j]) == 1))

morans_i <- function(x, W) {
  z <- x - mean(x)
  (length(x) / sum(W)) * sum(W * outer(z, z)) / sum(z^2)
}

set.seed(1)
x <- trend_strength * (row_of + col_of) / side + rnorm(n_areas)
observed_I <- morans_i(x, W)

permuted_I <- replicate(999, morans_i(sample(x), W))   # reshuffle values over locations
pseudo_p <- (sum(permuted_I >= observed_I) + 1) / (999 + 1)

hist(permuted_I, breaks = 30, col = "grey85", xlim = range(c(permuted_I, observed_I)),
     main = "Reference distribution under spatial randomness", xlab = "Moran's I")
abline(v = observed_I, col = "red", lwd = 2)
c(observed_I = round(observed_I, 3), expected_under_H0 = round(-1 / (n_areas - 1), 3),
  pseudo_p = pseudo_p)
```

## 2.12 The Moran scatter plot {#t2-12 .topic flag="objective"}

With **row-standardized** weights (S₀ = n), Moran's I simplifies to

$$
I = \frac{\sum_i z_i \cdot \left(\sum_j w_{ij} z_j\right)}{\sum_i z_i^2},
$$

which is the **slope of a regression line** of the spatial lag Σⱼwᵢⱼzⱼ on zᵢ. The **Moran scatter plot**:

- x-axis: each area's own (standardized) value;
- y-axis: its **spatially lagged** value (the average of its neighbours);
- slope = Moran's I.

The four quadrants classify areas:

| Quadrant | Own value | Neighbours | Type |
|---|---|---|---|
| Upper right | High | High | **High-High**: cluster (hot spot) |
| Lower left | Low | Low | **Low-Low**: cluster (cold spot) |
| Lower right | High | Low | **High-Low**: spatial outlier |
| Upper left | Low | High | **Low-High**: spatial outlier |

```{r}
#| label: fig-moran-scatter
#| echo: false
#| fig-width: 9
#| fig-height: 4
#| fig-cap: "Moran scatter plot for a simulated 10 × 10 map (row-standardized queen weights). Each point is one area: its own standardized value against the average of its neighbours. The slope of the line is Moran's I. Right: the quadrant of each area drawn back on the map. This is not yet a significance test; LISA (2.13) adds that."
side <- 10
x <- smooth_surface(side, 5, seed = 12) + rnorm(side^2, 0, 0.5)
W <- row_standardize(grid_weights(side, "queen"))
z <- as.vector(scale(x))
lag_z <- as.vector(W %*% z)
quad <- ifelse(z >= 0 & lag_z >= 0, 1, ifelse(z < 0 & lag_z < 0, 2, ifelse(z >= 0, 3, 4)))
qcols <- c("#d6604d", "#4393c3", "#f4a582", "#92c5de")   # HH, LL, HL, LH
par(mfrow = c(1, 2), mar = c(4.2, 4.2, 2.5, 1))
plot(z, lag_z, pch = 19, col = qcols[quad], xlab = "own value z (standardized)", ylab = "spatial lag: mean of neighbours' z",
     main = sprintf("Moran scatter plot: slope = I = %.2f", morans_i(x, W)))
abline(h = 0, v = 0, col = col_grey)
abline(lm(lag_z ~ z), lwd = 2)
usr <- par("usr")
text(usr[2], usr[4], "High-High", adj = c(1.05, 1.5), cex = 0.85, col = qcols[1], font = 2)
text(usr[1], usr[3], "Low-Low", adj = c(-0.1, -0.6), cex = 0.85, col = qcols[2], font = 2)
text(usr[2], usr[3], "High-Low", adj = c(1.05, -0.6), cex = 0.85, col = qcols[3], font = 2)
text(usr[1], usr[4], "Low-High", adj = c(-0.1, 1.5), cex = 0.85, col = qcols[4], font = 2)
par(mar = c(4.2, 1, 2.5, 1))
draw_grid(quad, side, qcols, main = "The same areas on the map")
legend("bottom", c("HH", "LL", "HL", "LH"), fill = qcols, horiz = TRUE, bty = "n", cex = 0.8, inset = -0.12, xpd = TRUE, border = NA)
```

"High" and "low" are **relative to the mean**.

**The slides' questions.** *Spatial clusters* (HH, LL): **positive** autocorrelation. *Spatial outliers* (HL, LH): **negative** autocorrelation (the area is unlike its neighbours).

## 2.13 Local spatial autocorrelation: LISA and local Moran's I {#t2-13 .topic flag="objective"}

| | Global | Local |
|---|---|---|
| Question | **Is** there spatial clustering? | **Where**, and what type? |
| Output | One number for the whole map | One statistic and one p-value **per area** |
| Measures | Moran's I (Geary's C) | **LISA**, for example local Moran's I; local Geary |
| Finds | Overall clustering | Hot spots, cold spots and **spatial outliers** |

**LISA** (local indicator of spatial association, Anselin 1995) must:

1. give a statistic for **each location** with an assessment of significance;
2. be such that the **sum of the local statistics is proportional to the global statistic**.

Generic local form Σⱼ wᵢⱼ f(xᵢ, xⱼ); generic global form Σᵢ Σⱼ wᵢⱼ f(xᵢ, xⱼ).

**Local Moran's I:**

$$
I_i = \frac{z_i \sum_j w_{ij} z_j}{\sum_i z_i^2 } ,
$$

the area's own deviation times the weighted sum of its neighbours' deviations. Global Moran's I corresponds to the **average** of the local statistics (their sum is proportional to the global I).

**Significance by conditional permutation.** As for the global test, except that **zᵢ is held fixed at location i** and only the *other* values are reshuffled over its neighbours. This gives a pseudo p-value for each area.

**Two maps** come out of a LISA analysis:

- a **significance map**: areas coloured by pseudo p-value (0.05, 0.01, 0.001);
- a **cluster map**: the significant areas coloured by type: **High-High** and **Low-Low** clusters, **High-Low** and **Low-High** outliers.

```{r}
#| label: fig-lisa
#| echo: false
#| fig-width: 9
#| fig-height: 3.6
#| fig-cap: "LISA, simulated. Left: the rate map. Middle: the significance map, pseudo p-values from conditional permutation. Right: the cluster map, the significant areas coloured by type. The one HL area is an odd high value placed inside a low region: a spatial outlier."
side <- 10
set.seed(31)
x <- smooth_surface(side, 5, seed = 12) + rnorm(side^2, 0, 0.5)
x[16] <- x[16] + 4                                        # one odd area inside a low region
W <- row_standardize(grid_weights(side, "queen"))
z <- as.vector(scale(x))
lag_z <- as.vector(W %*% z)
local_i <- z * lag_z
# conditional permutation: keep z_i fixed, reshuffle the other values over i's neighbours
n_perm <- 499
pseudo_p <- sapply(seq_along(z), function(i) {
  k <- sum(W[i, ] > 0)
  others <- z[-i]
  sims <- replicate(n_perm, z[i] * mean(sample(others, k)))
  (sum(abs(sims) >= abs(local_i[i])) + 1) / (n_perm + 1)
})
quad <- ifelse(z >= 0 & lag_z >= 0, 2, ifelse(z < 0 & lag_z < 0, 3, ifelse(z >= 0, 4, 5)))
cluster <- ifelse(pseudo_p <= 0.05, quad, 1)
sig_class <- ifelse(pseudo_p <= 0.001, 4, ifelse(pseudo_p <= 0.01, 3, ifelse(pseudo_p <= 0.05, 2, 1)))
par(mfrow = c(1, 3), mar = c(3.5, 0.5, 2.5, 0.5))
draw_grid(cut(x, quantile(x, 0:5 / 5), include.lowest = TRUE, labels = FALSE), side, seq_cols(5), main = "Rate (quintiles)")
draw_grid(sig_class, side, c("#f1f3f5", "#c7e9c0", "#74c476", "#238b45"), main = "Significance map")
legend("bottom", c("n.s.", "p ≤ 0.05", "p ≤ 0.01", "p ≤ 0.001"), fill = c("#f1f3f5", "#c7e9c0", "#74c476", "#238b45"),
       horiz = TRUE, bty = "n", cex = 1, inset = -0.1, xpd = TRUE, border = NA)
draw_grid(cluster, side, c("#f1f3f5", "#d6604d", "#4393c3", "#f4a582", "#92c5de"), main = "Cluster map")
legend("bottom", c("n.s.", "HH", "LL", "HL", "LH"), fill = c("#f1f3f5", "#d6604d", "#4393c3", "#f4a582", "#92c5de"),
       horiz = TRUE, bty = "n", cex = 1, inset = -0.1, xpd = TRUE, border = NA)
```

::: {.callout-warning}
## Interpreting a cluster map
"High" and "low" are relative to the **mean** of the variable. A High-High area is significant together with its neighbours: the cluster is the core area **and** its neighbours, not just the coloured area. With many areas, some will be significant at 0.05 by chance (multiple testing); stricter cut-offs (0.01, 0.001) or false discovery rate adjustments are often used. And LISA on raw rates inherits variance instability: small areas can produce spurious clusters.
:::

:::: {.quiz}
An area is significant in the LISA cluster map and coloured as **Low-High**. What does that mean?

- **Its value is below the mean while its neighbours' values are above it**
- Its value is below the mean and so are those of its neighbours
- Its value is above the mean while its neighbours' are below it
- It belongs to a cold spot that is surrounded by a hot spot

::: {.exp}
The first word describes the area itself, the second its neighbours (spatial lag), both relative to the mean. Low-High is a spatial outlier: a low value among high neighbours, which is negative local autocorrelation.
:::
::::

::: {.callout-tip}
## When to use global vs local statistics
Global Moran's I first, to ask whether there is clustering at all. LISA when you need to locate clusters and outliers, for example for surveillance or to target resources. Choose and report W, and check robustness to another W.
:::

## 2.14 Session summary and key questions {#t2-14 .topic}

```{mermaid}
%%| fig-cap: "Lecture map: how the session's problems lead into each other. Not causal."
flowchart LR
  A["Map counts"] -->|"population dominates"| B["Map crude rates"]
  B -->|"small areas unstable"| C["SMR / EB smoothing"]
  C --> D["Is there clustering?<br/>define W → global Moran's I"]
  D --> E["Where is it?<br/>LISA cluster map"]
```

::: {.callout-note collapse="true"}
## Why can a map of counts and a map of crude rates show different patterns?
Counts are spatially extensive: they grow with population, so the count map mostly shows where people live. Crude rates are intensive: they divide by population and show risk. A big city can have the most cases but an average rate; a small municipality can have few cases but a high rate.
:::

::: {.callout-note collapse="true"}
## Why does EB smoothing shrink small areas more than large ones?
The weight wᵢ = σ²/(σ² + μ/Pᵢ) compares true between-area variance with sampling noise μ/Pᵢ. In small areas the noise term is large, so wᵢ is small and the estimate is pulled toward the reference rate. In large areas the crude rate is already precise and keeps most of its weight.
:::

::: {.callout-note collapse="true"}
## What is the null hypothesis for Moran's I, and how does a permutation test use it?
Spatial randomness: values are unrelated to location, so any reshuffling of the values over the areas is equally likely. Reshuffling many times and computing I each time gives the distribution of I under H₀; the share of permuted values at least as large as the observed I is the pseudo p-value.
:::

::: {.callout-note collapse="true"}
## Can global Moran's I be close to zero even though there are local clusters?
Yes. Local positive (clusters) and negative (outliers) associations can cancel out, or a few strong clusters may be diluted by many random areas. That is why LISA is used to look locally.
:::
