---
title: "Session 1 — Introduction to Spatial Epidemiology"
subtitle: "Spatial data · file formats · coordinate reference systems · John Snow · what spatial epidemiology is for · theory · MAUP, ecological fallacy, dependence and heterogeneity"
engine: knitr
---

```{=html}
<script>document.body.classList.add("block-intro");</script>
```

```{r}
#| include: false
source("_helpers.R")
```

This first session sets up the vocabulary for the whole course. It answers three questions: **what makes data spatial**, **what spatial epidemiology is trying to achieve**, and **which problems are special to analysing health data by place**.

The thread running through the session, and the rest of the course, is a sentence from the last slide: *a map is often the beginning of the epidemiologic question, not the end.* Seeing **where** disease is high leads to asking **why** it is high there, and then **what can be done**.

The second half introduces the methodological issues that come back in every later session: the modifiable areal unit problem (MAUP), the ecological fallacy, spatial dependence and spatial heterogeneity.

::: {.callout-note collapse="true"}
## Sources and session
**Slides:** "session_1_v0-1.pdf" (Qinyun Lin, 6 October 2026, 13:15–15:00).

**In-class activity:** [the story map on John Snow and the 1854 cholera outbreak](https://docs.google.com/presentation/d/e/2PACX-1vRFzOI-TRtjj_B792rNnqTSShBAEmRsXXl2S795acpcYGRAX1755m6oH0ntNYTX5mG3eObawPuF0p5k/pub?start=false&loop=false&delayms=3000&slide=id.g798c99946b_0_331){target="_blank"} (Center for Spatial Data Science, University of Chicago).

**Lecture notes:** the topics marked ✏️ were written down in class as important. The notes also record that **the exam tests concepts, not programming**.
:::

::: {.callout-tip collapse="true"}
## Before you start: three ideas from earlier courses
**Rate vs count.** A count (20 cases) depends on how many people there are. A rate or risk (20 cases per 1,000) divides by the population at risk, so areas of different size can be compared.

**Confounding.** An association between an exposure and an outcome can be produced by a third factor related to both. Place is often a stand-in for many such factors at once (income, pollution, access to care).

**Independence in regression.** Ordinary least squares (OLS) and the usual standard errors assume that, after accounting for the covariates, observations are independent. Session 1 introduces the reason why spatial data often break this assumption: nearby places resemble each other.
:::

## 1.1 What is spatial data? {#t1-1 .topic flag="notes"}

**Spatial data = information + location.** The information (an attribute: a disease rate, an income, a pollution level) is tied to a place. The information may change with location; that is the point of studying it spatially. Very occasionally spatial data contain only location (a list of clinic coordinates), but **without location the data are no longer spatial**.

A spatial data file has two parts that belong together:

- the **geometry**: where each feature is (coordinates, or the outline of an area);
- the **attribute table**: one row per feature with its variables, like an ordinary data frame.

**Two ways of representing space**

| Model | How space is stored | Feature types | Typical health example |
|---|---|---|---|
| **Vector** | Exact coordinates of points and the vertices that join them | **Point** (one coordinate pair), **line** (connected points), **polygon / area** (closed line) | Case addresses (points), roads (lines), municipalities or DeSO areas (polygons) |
| **Raster** | A grid of equal cells; every cell holds a value | The same shapes, approximated by filled cells | Air-pollution surfaces, land cover, satellite temperature |

```{r}
#| label: fig-vector-raster
#| echo: false
#| fig-width: 9
#| fig-height: 4.6
#| fig-cap: "The three feature types in the two data models (after slide 5). Vector data store exact coordinates; raster data approximate the same shapes with filled grid cells."
par(mfrow = c(2, 3), mar = c(1, 1, 2.2, 1))
blank <- function(main) plot(NA, xlim = c(0, 10), ylim = c(0, 10), asp = 1, axes = FALSE, xlab = "", ylab = "", main = main)
# vector
blank("Vector: point"); box(col = col_grey)
points(c(2, 5, 7.5, 3.5), c(7, 4, 6.5, 2.5), pch = 19, col = col_orange, cex = 1.6)
text(5, 4, "case address", pos = 4, cex = 0.8)
blank("Vector: line"); box(col = col_grey)
road <- cbind(c(1, 3, 5, 6, 9), c(2, 4, 4.5, 7, 8.5))
lines(road, lwd = 3, col = col_blue); points(road, pch = 21, bg = "white", cex = 1.2)
text(6, 7, "road", pos = 4, cex = 0.8)
blank("Vector: polygon (area)"); box(col = col_grey)
poly <- cbind(c(1.5, 4, 8, 8.8, 6, 2), c(2, 1.2, 2.5, 7, 8.5, 6.5))
polygon(poly, col = "#2f5d8a33", border = col_blue, lwd = 2); points(poly, pch = 21, bg = "white")
text(5, 5, "municipality", cex = 0.8)
# raster
cells <- function(on, main) {
  blank(main)
  for (i in 0:9) for (j in 0:9) rect(i, j, i + 1, j + 1, col = if (on[i + 1, j + 1]) "#2f5d8aaa" else "white", border = "#dee2e6")
}
m <- matrix(FALSE, 10, 10); m[cbind(c(3, 6, 8, 4), c(8, 5, 7, 3))] <- TRUE; cells(m, "Raster: point")
m <- matrix(FALSE, 10, 10); m[cbind(c(2, 3, 4, 5, 6, 6, 7, 8, 9), c(3, 4, 5, 5, 6, 7, 8, 8, 9))] <- TRUE; cells(m, "Raster: line")
gx <- outer(1:10 - 0.5, rep(1, 10)); gy <- outer(rep(1, 10), 1:10 - 0.5)
inside <- matrix(sp_in <- mapply(function(x, y) {
  # point-in-polygon by ray casting
  n <- nrow(poly); j <- n; c <- FALSE
  for (i in 1:n) { if (((poly[i, 2] > y) != (poly[j, 2] > y)) &&
       (x < (poly[j, 1] - poly[i, 1]) * (y - poly[i, 2]) / (poly[j, 2] - poly[i, 2]) + poly[i, 1])) c <- !c; j <- i }
  c }, gx, gy), 10, 10)
cells(inside, "Raster: area")
```

In the lecture notes, *area = polygon*. Most of this course works with **areal (polygon) data**: counts or rates aggregated to administrative areas.

:::: {.quiz}
A study links each child's home address to the PM2.5 concentration of the 1 × 1 km cell it falls in. Which two data models are combined?

- **Vector points for homes and a raster grid for pollution**
- Vector polygons for homes and vector lines for pollution
- A raster grid for homes and vector points for pollution
- Two raster grids at different cell sizes for both layers

::: {.exp}
A home address is a single coordinate pair, a vector point. A pollution surface measured on 1 km cells is a raster. Joining them means looking up the cell value at each point.
:::
::::

## 1.2 Spatial file formats: CSV, shapefile, GeoJSON {#t1-2 .topic flag="notes"}

| Format | What it is | Watch out for |
|---|---|---|
| **CSV** (or other text file) | An ordinary table. It is spatial only if it has coordinate columns (longitude, latitude) | Cannot store polygons in a practical way; the CRS is not recorded in the file |
| **Shapefile** | ESRI's format. **At least four files** with the same name and different extensions | All files must sit **in the same folder**; if one is missing the layer breaks |
| **GeoJSON** | Open standard text format for points, lines and polygons with attributes | Can be large for detailed boundaries |

The four shapefile parts (as written down in class):

| Extension | Holds |
|---|---|
| `.shp` | The geometry: the shapes themselves |
| `.shx` | The index that lets software find each shape quickly |
| `.dbf` | The attribute table (one row per shape) |
| `.prj` | The projection: which coordinate reference system the coordinates are in |

::: {.misconception}
"The `.shp` file *is* the shapefile, so I only need to send that." The `.shp` holds the shapes but no attributes and no CRS. Without `.dbf` you lose the data columns, and without `.prj` the software does not know where on Earth the coordinates belong. Keep all parts together.
:::

:::: {.quiz}
A colleague emails you `kommuner.shp`, `kommuner.shx` and `kommuner.dbf`. The map opens, but distances come out strange and it will not line up with other layers. What is most likely missing?

- **The `.prj` file, so the coordinate reference system is unknown**
- The `.dbf` file, so the attribute columns cannot be read
- The `.shx` file, so the shapes cannot be indexed at all
- A GeoJSON copy, which shapefiles need for their geometry

::: {.exp}
Shapes and attributes load fine (`.shp`, `.shx`, `.dbf` are there). What is missing is the projection information in `.prj`: without it, software cannot place or transform the coordinates correctly.
:::
::::

## 1.3 Coordinate reference systems and EPSG codes {#t1-3 .topic flag="inferred"}

The Earth is round; a map is flat. A **coordinate reference system (CRS)** says how coordinates relate to places on the Earth's surface, and for a projected CRS, which method was used to **flatten (project)** the surface onto two dimensions. Different projections give visibly different world maps: every projection distorts area, shape, distance or direction somehow.

CRSs are identified by **EPSG codes** (from the European Petroleum Survey Group): a standard number that pins down the datum, the coordinate system and the projection.

| | EPSG:4326 (WGS 84) | EPSG:3006 (SWEREF 99 TM) |
|---|---|---|
| Kind | **Geographic** (unprojected) | **Projected** |
| Coordinates | Longitude and latitude | Easting and northing |
| Units | **Degrees** | **Metres** |
| Typical use | GPS, web maps, data downloaded online | Swedish national spatial data |
| Distances and areas | Not directly: a degree of longitude shrinks towards the poles | Yes, Euclidean distance in metres is meaningful |

Two practical rules from the slides:

1. **Recognise 4326.** Coordinates like `41.784766, -87.609299` (small numbers with many decimals, within ±90 and ±180) are almost certainly longitude/latitude in EPSG:4326. Try that first.
2. **Transform before you join.** When you combine layers, put everything in the **same CRS** first. Layers in different CRSs will not line up, and spatial joins or distances will be wrong.

For distance or area calculations, transform to a suitable projected CRS (in Sweden, SWEREF 99 TM). This returns in Session 2: Euclidean distance between points is only meaningful for projected coordinates.

:::: {.quiz}
You want to define neighbours as all health centres within 5 km of each other. Your coordinates are in EPSG:4326. What should you do first?

- **Transform the points to a projected CRS in metres, such as SWEREF 99 TM**
- Use a 5-degree distance band, because the coordinates are in degrees
- Nothing, since Euclidean distance works the same in any CRS
- Convert the shapefile to GeoJSON, which stores distances in metres

::: {.exp}
In EPSG:4326, coordinates are in degrees, and one degree is not a fixed number of metres. A 5 km threshold needs projected coordinates in metres (or great-circle distances). File format does not change the CRS.
:::
::::

## 1.4 Geocoding and Swedish small areas (DeSO) {#t1-4 .topic flag="notes"}

**Geocoding** turns an address into coordinates (longitude and latitude), so that individuals can be placed on a map or linked to the area they live in. Geocoding can **fail or be wrong** (missing or misspelled addresses, rural addresses placed at a postcode centroid). Such errors are rarely random: they are more common in some kinds of areas, which can bias a spatial analysis (see [1.13](#t1-13)).

Most health data are reported for **administrative areas**: province, county, postcode, census tract. In Sweden the smallest standard statistical unit is **DeSO** (*demografiska statistikområden*, demographic statistical areas), designed by Statistics Sweden to have roughly 700 to 2,700 inhabitants each, so that statistics can be published at fine scale without identifying people. The lecture notes record DeSO as *the smallest area in Sweden* used in practice.

Why the choice of unit matters is the subject of [1.10 Scale and MAUP](#t1-10).

## 1.5 John Snow and the 1854 cholera outbreak {#t1-5 .topic flag="inferred"}

The in-class activity: read [the story map](https://docs.google.com/presentation/d/e/2PACX-1vRFzOI-TRtjj_B792rNnqTSShBAEmRsXXl2S795acpcYGRAX1755m6oH0ntNYTX5mG3eObawPuF0p5k/pub?start=false&loop=false&delayms=3000&slide=id.g798c99946b_0_331){target="_blank"} (about 15 minutes) and answer three questions.

| Question | Answer |
|---|---|
| **Research question** | What causes cholera, and why did so many people in one part of Soho, London, die in a few days in 1854? |
| **Competing hypotheses** | **Miasma**: disease spreads through "bad air" from rotting matter (the dominant view). **Waterborne**: cholera is transmitted through contaminated water (Snow's hypothesis) |
| **Spatial data used** | The **locations of cholera deaths** (point data, stacked as bars at each house) and the **locations of public water pumps** on a street map. Snow also compared deaths by **which water company** supplied a household |

The map showed deaths clustered around the **Broad Street pump**. Exceptions were informative: workers at a brewery with its own water supply were largely spared, while some deaths far away were in people who preferred to fetch water from Broad Street. The pump handle was removed.

**Classifying Snow's work** with the describe–analyze–intervene framework ([1.7](#t1-7)): mapping the deaths **describes**; testing the water hypothesis against miasma (including the comparison of water companies) **analyzes**; removing the pump handle is an **intervention**. One study can move across all three goals, which is exactly what modern spatial epidemiology rarely does.

::: {.misconception}
"Snow's map proved the cause." The map alone shows clustering around a pump, which miasma could also explain (bad air near the pump). What made the argument strong was **theory** (a waterborne mechanism), **informative exceptions**, and a **comparison** between households with different water supply. A map starts the question; the comparison answers it.
:::

## 1.6 Definitions of spatial epidemiology {#t1-6 .topic flag="objective"}

Three definitions from the slides, from narrow to broad:

| Source | Definition (key words) |
|---|---|
| **Elliott & Wartenberg (2004)** | "the description and analysis of **geographic variations in disease** with respect to **demographic, environmental, behavioral, socioeconomic, genetic, and infectious risk factors**" |
| **Lawson (2018)** | uses geographic location to understand the **distribution of health outcomes and their determinants**, considering how **location influences the interaction** of risk factors |
| **Morrison et al. (2024)**, broader | applies theory and methods from **epidemiology, geography and statistics** to **describe** spatial distributions, **analyze** possible causes, **inform intervention**, and **improve health** |

The lecture notes record the first definition: *geographic variations in disease* with respect to demographic and other risk factors. Be able to reproduce one definition in your own words and say what the broader one adds: an explicit aim to explain and to intervene, not just to map.

## 1.7 Describe → analyze → intervene (Morrison et al. 2024) {#t1-7 .topic flag="inferred"}

Morrison et al. (2024, *Epidemiology* 35(4):542–555) reviewed the spatial epidemiology literature and asked whether the field is aligned with the broader goals of epidemiology.

```{mermaid}
%%| fig-cap: "Workflow: each arrow means \"then\", the next goal a study can move on to. Not causal. The three goals of epidemiology as used in Morrison et al. (2024)."
flowchart LR
  D["DESCRIBE<br/>Where is health distributed?<br/>(disease mapping, cluster detection,<br/>surveillance, resource allocation)"] --> A["ANALYZE<br/>Why might the pattern occur?<br/>(place-based determinants,<br/>exposure–outcome relationships,<br/>mechanisms, inequalities)"]
  A --> I["INTERVENE<br/>What can we change?<br/>(target interventions, evaluate<br/>place-based policies,<br/>change environments)"]
```

**What the review found** among 482 papers:

| Type | Number |
|---|---|
| Descriptive | 122 |
| Analytic | 167 |
| **Intervention** | **0** |
| Reviews | 76 |
| Demonstrations of methods | 117 |

**Zero intervention studies.** The field is good at mapping patterns and developing methods, but mainly describes and explains; it rarely evaluates what works. The slides' re-orientation: spatial statistical methods are **tools, not the endpoint**. The question is not only "Where are the clusters?" but "Why are they there, and what can we do about them?"

From *where* to *why* requires more than a map: **pattern → theory → analysis**. Theory tells you which spatial units, exposures, pathways, confounders and mechanisms are meaningful.

## 1.8 Why place matters for health {#t1-8 .topic flag="objective"}

Physical and social environments produce **health disparities**. Examples from the slides:

- longer distance to mammography facilities → later breast cancer diagnosis (Nattinger 2001);
- pedestrian-friendly environments → lower obesity (Gordon-Larsen 2006);
- living along major traffic corridors → cardiovascular disease (McEntee 2008).

Three frameworks that put place inside epidemiology:

| Framework | Core idea |
|---|---|
| **Social determinants of health** (WHO 2010) | The conditions in which people are born, grow, live, work and age, including **neighbourhood and physical environment** and **access to healthcare**, shape health |
| **Eco-epidemiology** (Schwartz et al. 1999) | Individuals are interdependent and connected to the biological, physical, social and historical **contexts** they live in |
| **Multiple exposures over the life span** (Buck Louis & Sundaram 2012) | Measure many environmental exposures at different times in life, not one exposure at one time |

The common message: place stands for **context**: access to food, care, education, green space and clean air.

## 1.9 Theoretical frameworks: social-ecological model and life course {#t1-9 .topic flag="notes"}

Both frameworks were written down in class.

**Social-ecological model** (Bronfenbrenner 1994; Dahlberg & Krug 2002). Health is shaped by nested levels:

| Level | Example from the slide (stigma around opioid use disorder) |
|---|---|
| Individual | Age, race/ethnicity, education |
| Interpersonal | Friends' or family members' experience with the disorder |
| Organizational | Workplaces, clinics, schools |
| **Community / neighbourhood** | **Spatial access** to treatment (medication for opioid use disorder) |

Spatial data enter at the outer level: the neighbourhood and what is reachable from it.

**Life-course perspective: five mechanisms behind social inequality in health** (Diderichsen et al. 2001):

| | Mechanism | Where spatial data can help |
|---|---|---|
| I | **Social stratification**: people end up in different social positions | Residential segregation by income |
| II | **Differential exposure**: lower positions meet more causes of illness | Pollution, noise, unhealthy food environments by area |
| III | **Differential vulnerability**: the same exposure harms some groups more | Area deprivation modifying the effect of heat |
| IV | **Differential disease consequences**: illness leads to worse outcomes for some | Distance to care, late diagnosis |
| V | **Consequences for the individual and society**: illness feeds back into social position | Job loss concentrated in some areas |

The in-class discussion question was *where spatial data and spatial epidemiology may be useful, and how*. A good answer links one mechanism to one spatial measure, as in the right-hand column.

## 1.10 Scale and the modifiable areal unit problem (MAUP) {#t1-10 .topic flag="objective"}

Health data are usually aggregated to **administrative zones** (province, county, postcode, census tract, DeSO). These units:

- are often **irregular and arbitrary**, drawn for administration, not for health;
- can **change over time** (inconsistent geography), which makes trends hard to follow.

**The modifiable areal unit problem (MAUP)** (Openshaw 1984): changing the spatial units can lead to **different statistical results and interpretations**, even when the underlying individual data are the same. The lecture notes put it as: *if we change scale (new areas are under investigation), the interpretation will change.*

MAUP has two parts:

| Effect | What changes | Example |
|---|---|---|
| **Scale effect** | The size of the units (DeSO → municipality → county) | A correlation between deprivation and mortality gets stronger as units get larger, because aggregation averages out individual noise |
| **Zoning effect** | Where the boundaries are drawn, at the same size | Two ways of drawing 20 equal-sized districts give different cluster maps |

```{r}
#| label: fig-maup-zoning
#| echo: false
#| fig-width: 9
#| fig-height: 3.3
#| fig-cap: "Zoning effect, simulated. The same 64 small units, with high values along a diagonal band, aggregated into four zones in three different ways. Vertical or horizontal strips hide the band almost completely; squares turn it into two 'hot' and two 'cold' quarters. Same data, three different maps."
side <- 8
rc <- grid_rc(side)
set.seed(3)
# 64 small units: high values along the diagonal band
unit_rate <- round(10 + 8 * exp(-((rc$row - rc$col)^2) / 6) + rnorm(64, 0, 1), 1)
zone_a <- (rc$col - 1) %/% 2                    # 4 vertical strips
zone_b <- (rc$row - 1) %/% 2                    # 4 horizontal strips
zone_c <- ((rc$row - 1) %/% 4) * 2 + (rc$col - 1) %/% 4   # 4 squares
br <- seq(9, 19, length.out = 6); cols <- seq_cols(5)
par(mfrow = c(1, 4), mar = c(1, 0.5, 3, 0.5))
draw_grid(unit_rate, side, cols, br, main = "64 small units")
for (z in list(list(zone_a, "Zoning 1: 4 vertical strips"), list(zone_b, "Zoning 2: 4 horizontal strips"),
               list(zone_c, "Zoning 3: 4 squares"))) {
  zone_mean <- ave(unit_rate, z[[1]])
  draw_grid(zone_mean, side, cols, br, main = z[[2]], border = NA)
  zr <- grid_rc(side)
  for (k in unique(z[[1]])) {
    cells <- which(z[[1]] == k)
    rect(min(zr$col[cells]) - 1, side - max(zr$row[cells]), max(zr$col[cells]), side - min(zr$row[cells]) + 1, border = "black", lwd = 1.5)
  }
}
```

A partial remedy from the slides: a **regular hexagon grid** (all units the same size and shape), but health data are rarely available on such grids.

```{r}
#| label: maup-demo
#| echo: false
#| fig-width: 9
#| fig-height: 3.4
#| fig-cap: "Scale effect, simulated. The same 3,600 individuals (exposure x and outcome y with a weak individual-level association, plus a smooth spatial trend in both) are aggregated to 900, 36 and 9 areas. The correlation of the area means grows as the areas get larger."
set.seed(530)
grid_side <- 60                                   # 60 × 60 individuals on a grid
row_pos <- rep(1:grid_side, times = grid_side)
col_pos <- rep(1:grid_side, each = grid_side)
area_trend <- (row_pos + col_pos) / grid_side      # smooth spatial gradient
exposure <- area_trend + rnorm(grid_side^2, sd = 1.2)
outcome  <- 0.2 * exposure + area_trend + rnorm(grid_side^2, sd = 1.5)

aggregate_to <- function(block_size) {
  area_id <- paste((row_pos - 1) %/% block_size, (col_pos - 1) %/% block_size)
  data.frame(x = tapply(exposure, area_id, mean),
             y = tapply(outcome, area_id, mean))
}
par(mfrow = c(1, 3), mar = c(4, 4, 3, 1))
for (block_size in c(2, 10, 20)) {
  area_means <- aggregate_to(block_size)
  plot(area_means$x, area_means$y, pch = 19, col = "#2f5d8a80",
       xlab = "mean exposure in area", ylab = "mean outcome in area",
       main = sprintf("%d areas: r = %.2f", nrow(area_means),
                      cor(area_means$x, area_means$y)))
}
```

The individual-level correlation in this simulation is about `r round(cor(exposure, outcome), 2)`. None of the area-level correlations is "the" answer: each describes associations **between areas at that scale**.

::: {.callout-tip}
## When to use this idea
Whenever results come from aggregated data: say which unit was used, ask whether conclusions hold at another scale, and prefer units that make sense for the mechanism (a health-care catchment for access to care, a smaller unit for air pollution).
:::

## 1.11 The ecological fallacy {#t1-11 .topic flag="objective"}

**Ecological fallacy** (Robinson 1950): drawing **incorrect inferences about individuals from group-level data**.

Example from the slides: a region has a high crime rate, so one concludes that an individual from that region is likely to commit a crime. The aggregate does not describe individual characteristics or behaviour.

```{r}
#| label: fig-eco-fallacy
#| echo: false
#| fig-width: 9
#| fig-height: 3.6
#| fig-cap: "Ecological fallacy, simulated. Left: across three areas, higher average exposure goes with a higher average outcome. Right: within every area, individuals with higher exposure have a lower outcome. The area-level association says nothing reliable about individuals."
set.seed(7)
area_names <- c("Area A", "Area B", "Area C")
area_cols <- c(col_blue, col_green, col_orange)
area_x_mean <- c(3, 5, 7)            # average exposure in each area
area_y_mean <- c(4, 6, 8)            # average outcome in each area
x <- y <- g <- NULL
for (k in 1:3) {
  xi <- rnorm(40, area_x_mean[k], 0.8)
  yi <- area_y_mean[k] - 0.8 * (xi - area_x_mean[k]) + rnorm(40, 0, 0.35)  # negative within each area
  x <- c(x, xi); y <- c(y, yi); g <- c(g, rep(k, 40))
}
par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))
plot(area_x_mean, area_y_mean, pch = 15, cex = 2.2, col = area_cols, xlim = c(1, 9), ylim = c(2, 10),
     xlab = "mean exposure in area", ylab = "mean outcome in area", main = "What the area data show")
abline(lm(area_y_mean ~ area_x_mean), lty = 2)
text(area_x_mean, area_y_mean, area_names, pos = 3, cex = 0.8)
mtext("positive association between areas", side = 3, line = 0.2, cex = 0.75)
plot(x, y, pch = 19, col = paste0(area_cols[g], "99"), xlim = c(1, 9), ylim = c(2, 10),
     xlab = "individual exposure", ylab = "individual outcome", main = "What happens to individuals")
for (k in 1:3) {
  xs <- seq(area_x_mean[k] - 1.6, area_x_mean[k] + 1.6, length.out = 2)
  lines(xs, area_y_mean[k] - 0.8 * (xs - area_x_mean[k]), col = area_cols[k], lwd = 2.5)
}
mtext("negative association within every area", side = 3, line = 0.2, cex = 0.75)
```

**MAUP vs ecological fallacy.** Both come from aggregation, but they ask different things:

| | MAUP | Ecological fallacy |
|---|---|---|
| The problem | Area-level results depend on **how the areas are drawn** | Area-level results are **wrongly applied to individuals** |
| Level of the conclusion | Stays at area level | Jumps from area to individual level |
| Question it raises | Would I see the same pattern with other units? | Does this association hold for people? |

:::: {.quiz}
Municipalities with more fast-food outlets per capita have higher diabetes prevalence. A newspaper writes: "People who eat fast food get diabetes." Which problem does this illustrate?

- **The ecological fallacy: an area association is applied to individuals**
- The modifiable areal unit problem: the boundaries were drawn badly
- Spatial heterogeneity: the association differs between regions
- Geocoding error: outlets were placed in the wrong municipality

::: {.exp}
The data are at municipality level. Whether the individuals with diabetes ate fast food is not observed; inferring it is the ecological fallacy. MAUP would be the issue if the conclusion were about municipalities but changed with other boundaries.
:::
::::

## 1.12 Spatial dependence and spatial heterogeneity {#t1-12 .topic flag="objective"}

These are the two properties that make spatial data special, and they motivate every method in the rest of the course.

**Spatial dependence.** *Tobler's first law of geography:* "things closer together are more similar than things further apart". Observations near one another may be more similar (or otherwise related) than observations far apart. Example: expensive and cheap neighbourhoods cluster because of socio-spatial processes. Dependence usually holds **within some distance**; beyond it, observations behave as independent.

Global measures summarise dependence in **one number**:

| Measure | Reading |
|---|---|
| **Moran's I** ≈ 0 | Little or no global spatial autocorrelation |
| **Moran's I** > 0 | Similar values tend to occur near one another (clustering) |
| **Moran's I** < 0 | Dissimilar values tend to occur near one another (checkerboard) |
| **Geary's C** | An alternative global measure (built on squared differences between neighbours) |

Both can be tested for statistical significance; Session 2 shows how.

**Spatial heterogeneity.** Processes and relationships **differ by location**, so one model applied everywhere can miss local patterns. Example: the relationship between income and housing prices differs between urban and rural areas. It shows up as regression coefficients that vary across regions, or as spatially structured random effects (Session 4).

| | Spatial dependence | Spatial heterogeneity |
|---|---|---|
| About | **Similarity between nearby values** | **Differences in processes or relationships across space** |
| Question | Are neighbours alike? | Is the same model valid everywhere? |
| Methods in this course | Moran's I, LISA (S2); spatial lag and error models (S3); structured random effects (S4) | Mentioned; GWR and similar models are outside the course (S3) |

:::: {.quiz}
The effect of temperature on hospital admissions is strong in the south of a country and weak in the north. Which concept is this?

- **Spatial heterogeneity: the relationship differs by location**
- Spatial dependence: neighbouring admission counts are similar
- The ecological fallacy: area data are applied to individuals
- The MAUP: the result depends on the boundaries used

::: {.exp}
A relationship (temperature → admissions) that changes across space is heterogeneity. Dependence would be about values in neighbouring areas resembling each other, which can be present at the same time but is a different property.
:::
::::

## 1.13 Other issues: reporting, geocoding, boundaries, sparse counts {#t1-13 .topic flag="objective"}

The slides list four further problems. Each is a form of **bias** in spatial data (the course syllabus asks you to reflect on bias from **aggregation, measurement and missing data**):

| Issue | Why it can bias a map or model |
|---|---|
| **Incomplete or geographically varying reporting** | If some regions report less completely, they look healthier than they are. The map shows the reporting system, not the disease |
| **Geocoding failure or error** | Failed addresses are dropped or misplaced; failures are often concentrated in rural or new areas (missing data that are not missing at random) |
| **Boundary changes over time** | Areas merge or split, so rates before and after are not comparable (inconsistent geography) |
| **Differences in population size / sparse counts** | Small areas give unstable rates: extreme values may be chance, not risk. Session 2 (smoothing) and Session 4 (Bayesian disease mapping) deal with this |

## 1.14 Tools: GIS, R and GeoDa {#t1-14 .topic flag="notes"}

**GIS (geographic information systems)** and related geospatial methods made spatial epidemiology practical: storing, linking, mapping and analysing data by location. Researchers use R, Stata, ArcGIS, GeoDa, Python and others.

In STA530:

- **R**: the main environment for spatial data analysis and modelling. Useful resources: [*Geocomputation with R*](https://geocompr.robinlovelace.net/){target="_blank"}, [*Spatial Data Science with R*](https://rspatial.org/){target="_blank"} and the [CSDS R spatial tutorials](https://spatialanalysis.github.io/tutorials/){target="_blank"}.
- **GeoDa**: free software for exploratory spatial data analysis and visualisation (maps, weights, Moran's I, LISA). **Install before 13 October** ([download GeoDa](https://geodacenter.github.io/download/){target="_blank"}).

The exam tests concepts, not software.

## 1.15 Session summary and key questions {#t1-15 .topic}

```{mermaid}
%%| fig-cap: "Lecture map: how the course questions lead into each other. Not causal."
flowchart LR
  W["WHERE?<br/>pattern"] --> Y["WHY?<br/>explanation / cause"] --> N["WHAT NEXT?<br/>intervention / action"]
```

**Why-questions** (answer in two or three sentences, then open the model answer):

::: {.callout-note collapse="true"}
## Why do you need the `.prj` file and a known CRS before joining two layers?
Because coordinates are only meaningful relative to a CRS. Two layers in different CRSs describe the same place with different numbers, so they will not overlay, and spatial joins or distance-based neighbours will be wrong. Transform both to the same CRS first, and use a projected CRS in metres for distances.
:::

::: {.callout-note collapse="true"}
## How do the MAUP and the ecological fallacy differ, given that both come from aggregation?
The MAUP is about area-level results changing when the areas are redrawn (different size or zoning), even with the same individuals. The ecological fallacy is about wrongly transferring an area-level association to individuals. A result can be robust to the MAUP and still be an ecological fallacy if applied to people.
:::

::: {.callout-note collapse="true"}
## What does "0 intervention studies" in Morrison et al. (2024) say about the field?
Spatial epidemiology has mainly described where disease is and analysed why, and has produced many methods, but has hardly ever evaluated whether place-based actions work. The re-orientation asks studies to go from map (describe) to mechanism (analyze) to action (intervene), as Snow's work did.
:::

::: {.callout-note collapse="true"}
## Why does spatial dependence matter for statistical analysis?
Most standard methods assume independent observations. If nearby areas resemble each other, there is less independent information than the number of areas suggests, and standard errors can be wrong. Dependence can also signal a spatially patterned omitted variable. Sessions 2–4 measure and model it.
:::
