---
title: "Session 4 — Introduction to Bayesian Disease Mapping"
subtitle: "The estimand and the likelihood · expected counts and SMR · partial pooling and shrinkage · unstructured and spatial random effects · BYM and BYM2 · posterior relative risk, credible intervals and exceedance probabilities"
engine: knitr
---

```{=html}
<script>document.body.classList.add("block-bayes");</script>
```

```{r}
#| include: false
source("_helpers.R")
```

Sessions 2 and 3 described and modelled spatial **dependence**. Session 4 shifts the focus to **estimating the underlying disease risk in each area**, especially when counts are small and crude rates are unstable.

The key question on slide 3: *when does a map reflect risk, and when does it reflect noise?* Two areas with the same SMR of 3.0 can carry very different amounts of information (3 observed vs 1 expected, or 300 vs 100).

The answer is a **Bayesian hierarchical model** that lets areas **borrow strength**: from all areas (partial pooling, unstructured effects) and from their **neighbours** (spatial smoothing, structured effects). The **BYM** and **BYM2** models combine the two. The output is not only a smoother map but a **posterior distribution** for every area's relative risk, which gives credible intervals and exceedance probabilities.

::: {.callout-note collapse="true"}
## Sources and session
**Slides:** "session_4_v0.pdf" (Qinyun Lin, 9 October 2026, 13:15–15:00).

**Learning objectives (Canvas post):** by the end of this session you should be able to

1. explain why crude small-area disease rates and standardized mortality ratios (SMRs) can be unstable;
2. recognize how the epidemiological outcome and denominator inform the choice of statistical likelihood;
3. explain the concepts of shrinkage, partial pooling and spatial smoothing;
4. describe the roles of spatially structured and unstructured random effects in BYM/BYM2 models;
5. interpret posterior relative risks, credible intervals and exceedance probabilities.

**Key questions to think about (Canvas post),** each answered in a section marked ❓:

1. Why might two geographic areas with the same SMR provide very different amounts of information about their underlying disease risks? → [4.1](#t4-1)
2. What is the difference between partial pooling and spatial smoothing? → [4.9](#t4-9)
3. Why does the BYM2 model include both spatially structured and unstructured variation? → [4.11](#t4-11)
4. Why might an area with a higher estimated relative risk have weaker evidence of elevated risk than another area? → [4.17](#t4-17)

**Optional reading:** [Gómez-Rubio, *Bayesian Inference with INLA*](https://becarioprecario.bitbucket.io/inla-gitbook/){target="_blank"}, a free online book on latent Gaussian models, spatial models and their implementation in R. Not required; you are not expected to understand the computational details of INLA.
:::

::: {.callout-tip collapse="true"}
## Before you start: three ideas you need
**Poisson regression with an offset.** For counts Y with exposure (person-time) T: log E[Y] = log T + Xβ. The offset log T has coefficient fixed at 1, so the model describes the rate Y/T.

**Bayes' theorem in words.** Posterior ∝ likelihood × prior: what you believe after the data combines what the data say with what you assumed before.

**Random effects (from mixed models).** Group-specific effects drawn from a common distribution, for example vᵢ ~ N(0, σ²). Groups with little data get estimates pulled toward the overall mean. Session 2's [Empirical Bayes smoothing](session2.qmd#t2-6) is a simpler version of the same idea.
:::

## 4.1 Same SMR, very different information {#t4-1 .topic flag="keyq"}

| | Area A | Area B |
|---|---|---|
| Observed Y | 3 | 300 |
| Expected E | 1 | 100 |
| SMR = Y/E | **3.0** | **3.0** |
| Uncertainty | **Wide** | **Much narrower** |

**Key question 1:** *why might two areas with the same SMR provide very different amounts of information?* Because the information depends on the **size of the counts**, not the ratio. With Y ~ Poisson(Eθ), the standard error of the SMR is roughly √Y / E = SMR/√Y. Area A: 3/√3 ≈ 1.7; area B: 3/√300 ≈ 0.17, ten times smaller. One or two extra cases in A would change its SMR enormously; in B they hardly matter.

A sensible model should let the **data-rich area speak more strongly** than the data-poor one.

```{r}
#| label: same-smr
#| echo: false
#| fig-width: 8
#| fig-height: 3
#| fig-cap: "Likelihood for the relative risk θ (scaled to a maximum of 1) in two areas with the same SMR = 3. Area A (3 observed, 1 expected) is compatible with almost anything from 1 to 8; area B (300 observed, 100 expected) pins θ down near 3."
theta <- seq(0.05, 9, length.out = 600)
lik_a <- dpois(3, 1 * theta);   lik_a <- lik_a / max(lik_a)
lik_b <- dpois(300, 100 * theta); lik_b <- lik_b / max(lik_b)
par(mar = c(4, 4, 1, 1))
plot(theta, lik_a, type = "l", lwd = 2, col = "#d9480f", ylim = c(0, 1),
     xlab = "relative risk θ", ylab = "relative likelihood")
lines(theta, lik_b, lwd = 2, col = "#2f5d8a")
abline(v = 3, lty = 2)
legend("topright", c("Area A: Y = 3, E = 1", "Area B: Y = 300, E = 100"),
       col = c("#d9480f", "#2f5d8a"), lwd = 2, bty = "n")
```

## 4.2 Start with the estimand: which likelihood? {#t4-2 .topic flag="objective"}

First ask **what is observed, what the denominator or reference is, and what quantity you want to estimate**.

| Outcome | Data | Likelihood | Target |
|---|---|---|---|
| **Incidence / mortality** | New events or deaths over population or **person-time** PTᵢ | Yᵢ ~ Poisson(PTᵢ λᵢ), i.e. log μᵢ = log PTᵢ + ηᵢ | **Rate** λᵢ |
| **Prevalence** | Existing cases among **Nᵢ individuals examined** | Yᵢ ~ Binomial(Nᵢ, pᵢ), logit(pᵢ) = ηᵢ | **Probability** pᵢ |
| **Disease mapping with expected counts** | Observed cases compared with an **expected count** Eᵢ | Yᵢ ~ Poisson(Eᵢ θᵢ) | **Relative risk** θᵢ |

ηᵢ is the linear predictor on the log (or logit) scale: possibly covariates plus spatial effects.

**For incidence**, the **offset** log PTᵢ fixes the amount of exposure; the model estimates the rate.

**Poisson or Binomial?** Both outcomes are counts. **The data structure decides the likelihood:**

| | Binomial | Poisson |
|---|---|---|
| Structure | Cases among a **fixed number N of individuals**, each yes/no | **Events over exposure** or person-time |
| Example (slide 8) | 20 people with obesity out of 100 examined | 15 deaths over 10,000 person-years |
| Target | Probability / prevalence pᵢ | Incidence or mortality rate λᵢ |

**A count of cases is not automatically Poisson.** And **BYM/BYM2 is a spatial random-effect structure, not intrinsically a Poisson model**: it can be added to the linear predictor of either.

<div class="drill" data-drill="likelihood"></div>

## 4.3 Observed and expected counts; indirect standardization {#t4-3 .topic flag="inferred"}

The classical disease-mapping model:

$$
Y_i \sim \text{Poisson}(E_i\,\theta_i)
$$

| Symbol | Meaning |
|---|---|
| Yᵢ | Observed number of cases or deaths in area i |
| Eᵢ | **Expected** cases if area i had the reference rates |
| θᵢ | **Relative risk**: θᵢ = 1 as expected; θᵢ > 1 elevated; θᵢ < 1 lower |

The inferential target is the **latent area-specific risk θᵢ**.

**Same model, two notations:**

| Disease-mapping notation | Regression notation |
|---|---|
| Yᵢ ~ Poisson(Eᵢ θᵢ), μᵢ = Eᵢ θᵢ | Yᵢ ~ Poisson(μᵢ), log μᵢ = log Eᵢ + ηᵢ, with ηᵢ = log θᵢ |

log Eᵢ is the **offset**.

**Where Eᵢ comes from: indirect standardization.**

$$
E_i = \sum_a n_{ia}\, r^*_a
$$

nᵢₐ is the population of area i in stratum a (for example age × sex) and r*ₐ the reference rate in that stratum. Eᵢ is the number of cases expected **from the area's own population composition**. This lets you compare areas with different age structures: an area with many elderly residents is not flagged just for being old.

<div class="drill" data-drill="smrstd"></div>

## 4.4 The SMR as the crude estimate {#t4-4 .topic flag="objective"}

$$
\text{SMR}_i = Y_i / E_i
$$

If each θᵢ is estimated **separately, with no smoothing**, the estimate is essentially the SMR (the maximum likelihood estimate of θᵢ under the Poisson model).

| Strength | Problem |
|---|---|
| Simple, transparent, easy to map | **Small denominators can produce extreme, unstable estimates** |

**Extreme ≠ necessarily high underlying risk.** An area with E = 0.5 and Y = 2 has SMR = 4, and with Y = 0 has SMR = 0, both from essentially no information. The variance of the SMR is roughly θᵢ/Eᵢ: the smaller Eᵢ, the noisier.

## 4.5 One Bayesian idea: combine data with a model for plausible risks {#t4-5 .topic flag="inferred"}

$$
\text{posterior} \;\propto\; \text{likelihood} \times \text{prior}
$$

| Part | Role in disease mapping |
|---|---|
| **Likelihood** | What the observed counts say about each area's risk |
| **Prior / latent model** | **How areas are allowed to share information** about their underlying risks |
| **Posterior** | Updated uncertainty about each area's underlying risk |

**The key modelling decision: how should information be shared across areas?** Not at all (SMR), with all areas equally (unstructured), or more with neighbours (spatial).

## 4.6 Partial pooling and shrinkage {#t4-6 .topic flag="objective"}

Extreme, noisy estimates move toward the common pattern.

::: {.worked}
| Area | Crude | Smoothed | Why |
|---|---|---|---|
| A | 3.00 | **1.45** | Little information → strong shrinkage |
| B | 3.00 | **2.75** | Much information → little shrinkage |
| C | 0.25 | **0.75** | Little information → pulled up toward the mean |

*Illustrative numbers from slide 15.*
:::

```{r}
#| label: fig-shrink-arrows
#| echo: false
#| fig-width: 9
#| fig-height: 3.2
#| fig-cap: "Shrinkage with the illustrative numbers from slide 15. Areas A and B have the same crude SMR of 3.0, but A has little data and is pulled far toward the overall level, while data-rich B keeps most of its signal. Area C's very low crude value is pulled up."
crude <- c(A = 3.00, B = 3.00, C = 0.25)
smoothed <- c(A = 1.45, B = 2.75, C = 0.75)
info <- c(A = "little data", B = "much data", C = "little data")
par(mar = c(4, 7, 2, 2))
plot(NA, xlim = c(0, 3.3), ylim = c(0.5, 3.5), yaxt = "n", xlab = "relative risk", ylab = "", main = "Crude → smoothed (slide 15, illustrative)")
axis(2, at = 3:1, labels = paste0("Area ", names(crude), "\n(", info, ")"), las = 1, cex.axis = 0.85, tick = FALSE)
abline(v = 1, lty = 2, col = col_grey)
for (k in 1:3) {
  y <- 4 - k
  arrows(crude[k], y, smoothed[k], y, length = 0.12, lwd = 2.5, col = if (info[k] == "much data") col_green else col_orange)
  points(crude[k], y, pch = 21, bg = "white", cex = 1.6)
  points(smoothed[k], y, pch = 19, col = col_blue, cex = 1.6)
}
legend("bottomright", c("crude SMR", "smoothed estimate"), pch = c(21, 19), pt.bg = "white", col = c("black", col_blue), bty = "n", cex = 0.8)
```

- **Less information → more shrinkage.**
- **More information → less shrinkage.**

**Shrinkage is not "making areas equal."** It weights noisy area-specific information against a shared model. A data-rich area with a real excess keeps it.

The three ways of pooling:

| | No pooling | **Partial pooling** | Complete pooling |
|---|---|---|---|
| Estimate | Each area alone (SMR) | Compromise, weighted by information | One common rate for all |
| Problem | Noisy | (the goal) | Ignores real differences |

**Try it.** Simulate 50 areas with true relative risks around 1, very different expected counts, and compare crude SMRs with a simple partial-pooling estimate (a Poisson–gamma Empirical Bayes estimate, the closed-form cousin of the Bayesian models below).

```{webr-r}
#| label: shrinkage
set.seed(4)
n_areas <- 50
expected <- round(exp(runif(n_areas, log(0.5), log(200))), 1)  # E_i: 0.5 to 200
true_rr  <- exp(rnorm(n_areas, 0, 0.25))                           # true theta_i
observed <- rpois(n_areas, expected * true_rr)
smr <- observed / expected

# Poisson-gamma partial pooling: prior theta ~ Gamma(a, b), posterior mean (a + Y) / (b + E)
prior_mean <- sum(observed) / sum(expected)
prior_var  <- max(var(smr) - mean(prior_mean / expected), 0.01)   # between-area variance
b <- prior_mean / prior_var
a <- prior_mean * b
smoothed <- (a + observed) / (b + expected)

par(mfrow = c(1, 2))
plot(expected, smr, log = "x", pch = 19, col = "#d9480f", ylim = range(c(smr, smoothed)),
     xlab = "expected count E (log scale)", ylab = "relative risk", main = "Crude SMR")
abline(h = 1, lty = 2)
plot(expected, smoothed, log = "x", pch = 19, col = "#2f5d8a", ylim = range(c(smr, smoothed)),
     xlab = "expected count E (log scale)", ylab = "relative risk", main = "Partially pooled")
abline(h = 1, lty = 2)
```

The smoothed estimates for small-E areas collapse toward the overall level; large-E areas barely move. Plots side by side make the comparison direct.

## 4.7 Unstructured heterogeneity: areas differ independently {#t4-7 .topic flag="objective"}

$$
\log \theta_i = \alpha + v_i, \qquad v_i \sim \text{Normal}(0, \sigma_v^2)
$$

| | |
|---|---|
| **What vᵢ captures** | Area-specific deviations from the overall level of risk α |
| **What it does not use** | **No neighbourhood information**: vᵢ and vⱼ are independent a priori |
| **Result** | A shared distribution → **partial pooling**: **shrinkage, but not spatial smoothing** |

Every area is shrunk toward the **same global level**, regardless of where it is. This is the fully Bayesian version of Session 2's Empirical Bayes idea.

## 4.8 Spatial smoothing: the CAR / ICAR prior {#t4-8 .topic flag="objective"}

Geography matters: neighbouring areas may share latent risk because they share

- **environment** (air pollution, built environment, climate);
- **social context** (deprivation, housing, local norms);
- **healthcare** (access, catchments, diagnostic practice);
- **unmeasured factors** (spatially patterned determinants not in X).

A **spatial prior** lets neighbouring areas share information about their underlying risks. In the conditional autoregressive form, **an area's latent effect is centred on its neighbours**:

$$
u_i \mid u_{-i} \;\sim\; \text{Normal}\!\left(\frac{\sum_j w_{ij} u_j}{n_i},\; \frac{\sigma_u^2}{n_i}\right)
$$

| Part | Meaning |
|---|---|
| **Neighbour mean** Σⱼwᵢⱼuⱼ / nᵢ | The conditional centre is set by adjacent areas |
| **nᵢ** | Number (or total weight) of neighbours |
| **Conditional variance** σᵤ²/nᵢ | **More neighbours → greater conditional precision** (under the standard intrinsic CAR, ICAR, specification) |

**Intuition (slide 18).** If A and C have similar spatial effects, the prior encourages B, between them, to be similar, **unless B's own data strongly disagree**. This is **borrowing strength through geography**.

For this course: understand the mechanism; you do not need to derive the joint distribution.

## 4.9 Partial pooling vs spatial smoothing {#t4-9 .topic flag="keyq"}

**Key question 2:** *what is the difference between partial pooling and spatial smoothing?*

| | Partial pooling (unstructured vᵢ) | Spatial smoothing (structured uᵢ) |
|---|---|---|
| Shrink toward | The **global** mean of all areas | The **local** mean of each area's **neighbours** |
| Uses the map? | **No**: shuffling the areas on the map changes nothing | **Yes**: the neighbourhood graph W decides who borrows from whom |
| Captures | Area-specific, non-spatial heterogeneity | Smooth, clustered geographic patterns |
| Model | Hierarchical model with exchangeable effects | CAR / ICAR prior |

```{r}
#| label: fig-smoothing-maps
#| echo: false
#| fig-width: 9
#| fig-height: 3.4
#| fig-cap: "Simulated map of 100 small areas (expected counts 0.5 to 15). The raw SMR map is speckled with noise. Unstructured pooling shrinks every area toward the global mean, so noise is reduced but the map ignores geography. Spatial smoothing pulls each area toward its neighbours and recovers the true pattern best. Illustrative Poisson–gamma estimates, not a fitted BYM2 model."
side <- 10
W <- grid_weights(side, "queen")
set.seed(14)
true_log_rr <- 0.5 * smooth_surface(side, 6, seed = 14)
expected <- round(exp(runif(side^2, log(0.5), log(15))), 1)     # many small areas
observed <- rpois(side^2, expected * exp(true_log_rr))
smr <- observed / expected
# unstructured partial pooling: Poisson-gamma toward the global mean
global_mean <- sum(observed) / sum(expected)
between_var <- max(var(smr) - mean(global_mean / expected), 0.02)
b <- global_mean / between_var; a <- global_mean * b
pooled <- (a + observed) / (b + expected)
# spatial smoothing (illustrative): the same, but each area's prior mean is its neighbourhood's pooled rate
local_mean <- as.vector((W %*% observed) / (W %*% expected))
spatial <- (local_mean * b + observed) / (b + expected)
br <- c(0, 0.6, 0.8, 0.95, 1.05, 1.25, 1.6, Inf); cols <- div_cols(7)
par(mfrow = c(1, 4), mar = c(0.5, 0.5, 3, 0.5), oma = c(2.5, 0, 0, 0))
draw_grid(exp(true_log_rr), side, cols, br, main = "True relative risk\n(unknown in practice)")
draw_grid(smr, side, cols, br, main = "Raw SMR\nno pooling")
draw_grid(pooled, side, cols, br, main = "Unstructured\nshrink to global mean")
draw_grid(spatial, side, cols, br, main = "Spatial\nshrink to neighbours")
bottom_legend(c("< 0.6", "0.6–0.8", "0.8–0.95", "≈ 1", "1.05–1.25", "1.25–1.6", "> 1.6"), cols)
```

Both shrink noisy small-area estimates. **Spatial smoothing is a particular kind of partial pooling** where the "group" an area borrows from is its neighbourhood.

## 4.10 The neighbourhood graph returns, in a different role {#t4-10 .topic flag="inferred"}

| Earlier: Moran's I, SAR, SEM | Here: Bayesian disease mapping |
|---|---|
| W summarises **observed** neighbouring values, or links **outcomes / errors** through a spatial process | Adjacency defines which **latent** area effects are **encouraged to be similar** |
| Spatial lag or error structure | A **spatial prior** for latent effects |

**Same neighbourhood graph, different role in the statistical model.**

## 4.11 The Besag–York–Mollié (BYM) model: why both uᵢ and vᵢ {#t4-11 .topic flag="objective"}

$$
Y_i \sim \text{Poisson}(E_i \theta_i), \qquad \log \theta_i = \alpha + u_i + v_i
$$

| Term | Role |
|---|---|
| **α**, overall level | Baseline log relative risk |
| **uᵢ**, **structured** | Spatially correlated component (ICAR): neighbouring areas **borrow strength** |
| **vᵢ**, **unstructured** | **Independent** area-specific heterogeneity |

**Two kinds of residual variation: spatial pattern + local noise.** The model is named after [Besag, York & Mollié (1991)](https://doi.org/10.1007/BF00116466){target="_blank"} (original paper, not required). Not all between-area variation needs to be spatially structured.

**Key question 3:** *why include both structured and unstructured variation?*

- **Structured uᵢ** captures smooth or clustered geographic pattern: if one area is high, its neighbours are compatible with being high.
- **Unstructured vᵢ** captures area-specific departures that **do not need to resemble neighbours**: local heterogeneity remains possible.
- **Without vᵢ**, every residual difference is pushed into the spatial surface: a genuinely odd area would be forced to look like its neighbours (over-smoothing), and its neighbours distorted toward it.
- **Without uᵢ**, geography is ignored: only global shrinkage.

Including both lets the **data decide** how much of the extra variation is spatial.

Both components represent **unexplained** variation in underlying risk; the distinction between them is **statistical**, not a claim about causes.

## 4.12 BYM2: separating overall variability from the spatial fraction {#t4-12 .topic flag="objective"}

**The problem with classic BYM** (slide 23):

- the marginal variability of uᵢ depends on the **entire neighbourhood graph**;
- unlike vᵢ, its variance parameter does **not** directly represent marginal variability;
- so how much residual variation is spatially structured is **not directly interpretable**, and priors are hard to set.

**BYM2** reparameterizes the same idea:

$$
\log \theta_i = \alpha + \frac{1}{\sqrt{\tau}}\left(\sqrt{\rho}\; u_i^{*} + \sqrt{1 - \rho}\; v_i\right)
$$

| Parameter | Meaning |
|---|---|
| **τ**, overall precision | 1/√τ controls the **overall magnitude** of residual heterogeneity. **Larger τ → less** residual variability; **smaller τ → more** |
| **ρ**, spatial fraction (mixing parameter), 0 ≤ ρ ≤ 1 | ρ close to 1 → **mostly structured**; ρ close to 0 → **mostly unstructured** |
| **uᵢ\***, scaled spatial effect | Scaled so its variance has a **stable interpretation across graphs** (marginal variance about 1, like vᵢ) |

This is the parameterization commonly used in modern Bayesian disease mapping. Original paper (not required): [Riebler et al. (2016)](https://doi.org/10.1177/0962280216660421){target="_blank"}.

**Interpreting ρ:**

| ρ ≈ 0 | ρ ≈ 0.5 | ρ ≈ 1 |
|---|---|---|
| Residual heterogeneity is predominantly **unstructured** | **Both** components matter | Residual heterogeneity is dominated by **geographic structure** |

```{r}
#| label: fig-bym2-rho
#| echo: false
#| fig-width: 9
#| fig-height: 3.4
#| fig-cap: "What the BYM2 mixing parameter ρ does, simulated on a 12 × 12 grid. The same structured component u* (an ICAR draw) and unstructured component v, mixed as √ρ·u* + √(1−ρ)·v. With ρ = 0 the residual pattern is pure speckle; with ρ = 1 it is a smooth surface where neighbours resemble each other."
side <- 12
W <- grid_weights(side, "rook")
Q <- diag(rowSums(W)) - W                       # ICAR precision
eig <- eigen(Q, symmetric = TRUE)
keep <- eig$values > 1e-8                       # drop the constant direction
set.seed(2)
z <- rnorm(sum(keep))
u <- as.vector(eig$vectors[, keep] %*% (z / sqrt(eig$values[keep])))
u_star <- u / sqrt(exp(mean(log(diag(MASS::ginv(Q))))))  # scale to geometric-mean variance 1
u_star <- as.vector(scale(u_star))
v <- rnorm(side^2)
br <- c(-Inf, -1.2, -0.6, -0.2, 0.2, 0.6, 1.2, Inf); cols <- div_cols(7)
par(mfrow = c(1, 3), mar = c(1, 0.5, 3, 0.5))
for (rho in c(0, 0.5, 1)) {
  effect <- sqrt(rho) * u_star + sqrt(1 - rho) * v
  draw_grid(effect, side, cols, br, main = sprintf("ρ = %s: %s", rho, c("all unstructured", "half and half", "all spatial")[match(rho, c(0, 0.5, 1))]))
}
```

::: {.misconception}
**The slide's question:** if ρ is estimated as 0.9, does it mean 90% of disease cases are caused by spatial factors? **No.** ρ describes the **residual latent variation** (what remains after α and any covariates) on the BYM2 mixing scale: about 90% of the *unexplained between-area variance* is spatially structured. It says nothing about causes, nothing about cases, and nothing about variation that the covariates already explain. Also, BYM2's ρ is not SAR's ρ ([3.4](session3.qmd#t3-4)).
:::

:::: {.quiz}
A BYM2 fit gives τ large and ρ ≈ 0.85. What is the best summary?

- **Little residual variation overall, and most of it is spatially structured**
- Much residual variation overall, and most of it is unstructured noise
- About 85% of all cases in the region are caused by spatial factors
- The covariates explain 85% of the variation between the areas

::: {.exp}
τ is a precision, so a large τ means small residual variability. ρ near 1 means the residual variation that exists is mostly structured. ρ is not a causal share of cases and not an R².
:::
::::

## 4.13 Priors: what are we putting priors on? {#t4-13 .topic flag="inferred"}

| Part | Prior |
|---|---|
| **Fixed effects** α, β | Usually weakly informative or context-appropriate |
| **Overall variability** | Prior on τ (or the residual standard deviation 1/√τ) |
| **Spatial mixing** | Prior on ρ: expresses assumptions about the balance between structured and unstructured variation |

For an introductory course, the key point is that **priors regularize complexity**. In applied work, consider **prior sensitivity**.

- A prior on the **overall variability** influences **how much** smoothing occurs.
- A prior on the **spatial mixing** influences the **type** of smoothing.

## 4.14 Penalized-complexity (PC) priors {#t4-14 .topic flag="optional"}

Marked **optional depth** on the slide.

**Idea:** start from a simpler **base model** (for BYM2: no residual variation, or no spatial structure) and **penalise departures** from it, so complexity is only used when the data support it. Original paper (not required): [Simpson et al. (2017)](https://doi.org/10.1214/16-STS576){target="_blank"}.

**Why useful:** hyperpriors can be stated as **interpretable probability statements** about plausible effect sizes, for example "P(residual SD > 1) = 0.01" or "P(ρ < 0.5) = 2/3", rather than arbitrary values for a precision parameter. This is a common strategy with INLA and BYM2.

## 4.15 Adding area-level covariates {#t4-15 .topic flag="inferred"}

$$
\log \theta_i = \alpha + X_i\beta + u_i + v_i
$$

| | |
|---|---|
| **β** | Area-level association with relative risk, on the log scale |
| **exp(β)** | **Relative-risk ratio** for a one-unit difference in X, conditional on the model |
| **Caution** | Spatial random effects **do not automatically solve confounding or make β causal** |

Disease mapping and **ecological regression** overlap, but the inferential question matters: estimating each area's risk (mapping) or estimating an area-level association (regression).

**Spatial confounding.** If X itself is spatially smooth, X and the spatial effect uᵢ can explain the same geographic pattern; adding uᵢ can then change β a lot, and neither version is automatically "right".

<div class="drill" data-drill="logrr"></div>

## 4.16 What the posterior gives: relative risks, credible intervals, exceedance {#t4-16 .topic flag="objective"}

| Output | What it is | Answers |
|---|---|---|
| **Posterior mean / median** of θᵢ | A **smoothed point estimate**; often mapped as relative risk | How high is the risk? |
| **Credible interval** (CrI) | Range of posterior uncertainty for θᵢ | How sure are we? |
| **Exceedance probability** P(θᵢ > c ∣ data) | Probability that θᵢ exceeds a scientifically meaningful threshold c | How likely is it that risk is elevated? |

**Bayesian disease mapping lets you map uncertainty, not only point estimates.**

**Interpreting a 95% credible interval.** Example: posterior 95% CrI for θᵢ = [1.08, 1.42]. *Under the fitted model and priors, 95% of the posterior probability for this area's (latent) relative risk lies between 1.08 and 1.42.*

| A credible interval is | It is not |
|---|---|
| A **probability statement about the parameter**, given the data, model and prior | "95% of repeated confidence intervals contain the fixed true value" (the frequentist confidence interval statement) |
| **Conditional** on the model, data and prior assumptions | A statement free of assumptions |

**Exceedance probabilities.**

| Threshold | Meaning |
|---|---|
| c = 1 | Probability that risk is **above the reference level** |
| c = 1.2 | Probability that risk exceeds a **20% elevation** |

The choice of c should be tied to the **scientific or decision context**, not chosen to make a map look interesting. A map of exceedance probabilities can be more informative than thresholding a map of point estimates. A common convention flags areas with P(θᵢ > 1) above 0.8 or 0.95.

## 4.17 Higher posterior RR does not always mean stronger evidence {#t4-17 .topic flag="keyq"}

| | Area A | Area B |
|---|---|---|
| Posterior mean RR | **1.30** | 1.20 |
| 95% CrI | 0.75–2.00 | 1.10–1.31 |
| P(RR > 1) | 0.72 | **0.995** |

```{r}
#| label: fig-posterior-ab
#| echo: false
#| fig-width: 9
#| fig-height: 3.4
#| fig-cap: "Posterior distributions for the two areas on slide 32, approximated by log-normal curves matching the slide's 95% credible intervals. Area A's distribution is wide, so a large share lies below 1. Area B's is narrow and almost entirely above 1. The shaded area is the exceedance probability P(θ > 1); with this approximation A's is 0.79 rather than the slide's 0.72."
# lognormal approximations with the slide's means and intervals
area_a <- c(meanlog = log(1.22), sdlog = (log(2.00) - log(0.75)) / (2 * 1.96))
area_b <- c(meanlog = log(1.20), sdlog = (log(1.31) - log(1.10)) / (2 * 1.96))
rr <- seq(0.3, 3, length.out = 600)
da <- dlnorm(rr, area_a[1], area_a[2]); db <- dlnorm(rr, area_b[1], area_b[2])
par(mfrow = c(1, 2), mar = c(4.2, 3, 3, 1))
for (k in 1:2) {
  d <- list(da, db)[[k]]; prm <- list(area_a, area_b)[[k]]
  plot(rr, d, type = "l", lwd = 2, col = col_blue, yaxt = "n", ylab = "", xlab = "relative risk θ",
       main = c("Area A: high mean, wide", "Area B: lower mean, narrow")[k], ylim = c(0, max(db)))
  polygon(c(rr[rr > 1], rev(rr[rr > 1])), c(d[rr > 1], rep(0, sum(rr > 1))), col = "#d9480f55", border = NA)
  abline(v = 1, lty = 2)
  mtext(sprintf("shaded area = P(θ > 1) ≈ %.2f", 1 - plnorm(1, prm[1], prm[2])), side = 3, line = 0.1, cex = 0.8)
}
```

**Which area gives stronger evidence of elevated risk?** **Area B.** A has the higher point estimate, but its interval includes 1 and P(RR > 1) is only 0.72. B's estimate is lower but precise, entirely above 1.

**Key question 4:** *why might an area with a higher estimated RR have weaker evidence of elevated risk?* Because **magnitude and certainty are different dimensions**. A small area (few expected cases) has a wide posterior: even after shrinkage its mean can be high while the data are compatible with no excess. A large area can have a modest but precisely estimated excess.

<div class="drill" data-drill="exceed"></div>

## 4.18 Raw SMR → non-spatial smoothing → spatial smoothing {#t4-18 .topic flag="inferred"}

| | Raw SMR | Unstructured model | BYM / BYM2 |
|---|---|---|---|
| Each area | **Stands alone** | **Partially pools** toward an overall distribution | Partial pooling **+ neighbouring areas inform the structured component** |
| Sparse counts | Very sensitive | Stabilised | Stabilised |
| Uses geography | No | **No** | **Yes, explicitly** |
| Uncertainty | Only frequentist CI per area | Posterior for each area | Posterior for each area |

**The gain is not merely a prettier map: it is a model-based estimate of latent risk with uncertainty.** Unstructured hierarchical models borrow through a shared distribution; BYM2 additionally uses spatial relationships.

## 4.19 Cautions: smoothing can hide real boundaries; islands {#t4-19 .topic flag="inferred"}

**Spatial smoothing can hide real boundaries** (slide 34):

- administrative neighbours **need not share the same underlying process**;
- a **river, road, policy boundary or healthcare catchment** may create a real discontinuity;
- strong smoothing can **attenuate sharp local signals** (a single polluted area, a point source);
- so the **adjacency definition is a substantive modelling choice**.

Borrowing strength helps **when the borrowing assumption is plausible**.

**Islands / areas with no neighbours** (slide 35). An ICAR component is defined through the neighbourhood graph, so an isolated area has **no adjacency information to borrow from**. Handling: check the graph carefully; define meaningful connections when scientifically justified (for example a ferry link); otherwise rely on the unstructured component or the software's handling. **Always inspect the neighbourhood graph before fitting.** (Compare isolates in [Session 2](session2.qmd#t2-9).)

## 4.20 Workflow and model checking {#t4-20 .topic flag="inferred"}

```{mermaid}
%%| fig-cap: "Workflow: each arrow means \"then\". Not causal. A practical Bayesian disease-mapping analysis (slide 37)."
flowchart LR
  A["1 · Define outcome<br/>incidence? prevalence?<br/>observed/expected?"] --> B["2 · Build graph<br/>meaningful<br/>neighbourhood structure"]
  B --> C["3 · Fit model<br/>Poisson/Binomial + BYM2;<br/>consider covariates"]
  C --> D["4 · Check<br/>sensitivity, fit, residual<br/>pattern, prior influence"]
  D --> E["5 · Communicate<br/>map risk AND uncertainty;<br/>interpret cautiously"]
```

**What to check after fitting** (slide 38):

| Check | Question |
|---|---|
| **Posterior predictive fit** | Can the fitted model reproduce relevant features of the observed data? |
| **Residual structure** | Is meaningful spatial pattern still unexplained? |
| **Sensitivity** | Do conclusions change with priors, graph choices or model specification? |
| **Complexity** | Do extra components improve inference enough to justify them? |

**Posterior maps are the beginning of interpretation, not the end of model checking.** A model is one part of the epidemiological argument, not the entire argument.

## 4.21 Why INLA is common {#t4-21 .topic flag="optional"}

You are **not expected to understand the computational details** of INLA (Canvas post).

| | |
|---|---|
| **Latent Gaussian models** | BYM/BYM2 fit naturally in this framework |
| **Fast approximation** | INLA (integrated nested Laplace approximation) approximates posterior marginals **without running conventional MCMC** |
| **Practical workflow** | R-INLA supports adjacency graphs, BYM2, priors, posterior summaries and predictions |

INLA is a **computational approach**: it does not change the model, only how the posterior is approximated. Original paper (not required): [Rue, Martino & Chopin (2009)](https://doi.org/10.1111/j.1467-9868.2008.00700.x){target="_blank"}.

Optional reading (Canvas post): [Gómez-Rubio, *Bayesian Inference with INLA*](https://becarioprecario.bitbucket.io/inla-gitbook/){target="_blank"}. Not required reading.

## 4.22 A high-risk area is not a causal explanation {#t4-22 .topic flag="inferred"}

| Caution | Why |
|---|---|
| **Ecological inference** | Area-level associations need not represent individual-level relationships (ecological fallacy, [1.11](session1.qmd#t1-11)) |
| **Spatial confounding** | Covariates and latent spatial effects can share similar geographic patterns |
| **MAUP / boundaries** | Results depend on how areas are defined and aggregated ([1.10](session1.qmd#t1-10)) |

**"This area has elevated posterior risk" ≠ "living here causes the disease."**

**Spatial regression vs disease mapping** ask related but different questions:

| Spatial regression (Session 3) | Bayesian disease mapping (Session 4) |
|---|---|
| How should regression change when observations are spatially dependent? | How can we estimate underlying area risk when counts are noisy and geographically structured? |
| Focus: **association, dependence, residual structure** | Focus: **latent risk, smoothing, uncertainty** |

Both require you to make the spatial mechanism explicit.

## 4.23 Checkpoint: reconstruct the logic {#t4-23 .topic flag="keyq"}

Slide 41 asks you to explain each arrow in one sentence. Try it before opening the answer.

```{mermaid}
%%| fig-cap: "Lecture map: each arrow is one step of the argument. Not causal."
flowchart LR
  A["Observed /<br/>expected"] --> B["SMR"] --> C["Poisson<br/>model"] --> D["Partial<br/>pooling"] --> E["CAR /<br/>ICAR"] --> F["BYM2"] --> G["Posterior risk<br/>+ uncertainty"]
```

::: {.callout-note collapse="true"}
## Model answer: one sentence per arrow
1. **Observed/expected → SMR:** dividing observed by expected cases (from indirect standardization) gives a crude relative risk that adjusts for population composition.
2. **SMR → Poisson model:** treating Yᵢ ~ Poisson(Eᵢθᵢ) makes the SMR the maximum likelihood estimate of θᵢ and shows that its uncertainty depends on the size of the counts.
3. **Poisson model → partial pooling:** giving the log relative risks a common distribution lets data-poor areas borrow from the others, shrinking unstable SMRs toward the overall level.
4. **Partial pooling → CAR/ICAR:** a spatial prior centres each area's effect on its neighbours' effects, so borrowing happens locally (spatial smoothing).
5. **CAR/ICAR → BYM2:** combining a scaled structured effect with an unstructured effect, mixed by ρ and scaled by τ, lets the data decide how much residual variation is spatial.
6. **BYM2 → posterior risk + uncertainty:** the fitted model gives a posterior for each θᵢ, summarised as posterior means, credible intervals and exceedance probabilities.
:::

## 4.24 Take-home messages and key questions {#t4-24 .topic}

1. **Start with the estimand.** Incidence, prevalence and standardized burden imply different data structures.
2. **Raw small-area rates are noisy.** SMRs can be extreme simply because denominators are small.
3. **Bayesian models partially pool.** Data-poor areas shrink more; data-rich areas keep more of their own signal.
4. **BYM2 separates two ideas:** overall heterogeneity and the fraction that is spatially structured.
5. **Map uncertainty too.** Posterior RR, credible intervals and exceedance probabilities answer different questions.

::: {.callout-note collapse="true"}
## Why is Empirical Bayes (Session 2) not the same as BYM2?
EB estimates the prior from the data and shrinks every area toward one global rate, giving point estimates only. BYM2 is fully Bayesian: it puts priors on the hyperparameters, gives a full posterior for every area, and includes a spatial component, so areas are also smoothed toward their neighbours.
:::

::: {.callout-note collapse="true"}
## When could spatial smoothing give a misleading map?
When neighbouring areas do not share the same underlying process: a real boundary (river, motorway, catchment border, policy line) or a sharp local hazard. The ICAR prior pulls the high area down and its neighbours up, blurring a genuine signal. Check the adjacency definition and compare with a less smoothed model.
:::

::: {.callout-note collapse="true"}
## Prevalence of diabetes from a health survey in each region: Poisson or Binomial?
Binomial: Yᵢ people with diabetes out of Nᵢ respondents, each yes/no, with target pᵢ (logit link). A BYM2 term can still be added to logit(pᵢ). Poisson would fit new diagnoses over person-time.
:::
